---
title: "Data_cleaning"
output: html_document
date: "2025-09-21"
---

```{r library}
library(here)
library(readr)
library(tidyverse)
```
others data
```{r set up file path(relative)}
rm(list = ls())
wd <- "analysis/pnwandmore/data/other_treerings/direct_searching/others/doi_10_5061_dryad_k5k9074__v20190226/Raw_Ring_Widths_012118/Raw_Ring_Widths_012118"
wo <- 'analysis/pnwandmore/input/other_tree_rings/direct_searching/others'
wl <- "analysis/pnwandmore/data/other_treerings/direct_searching/others/doi_10_5061_dryad_k5k9074__v20190226/Tree_Info_BAI_012118/Tree_Info_BAI_012118"
```

```{r pressure, echo=FALSE}
rawRingDatas <- list.files(here(wd),pattern="avgRingWidth.*",recursive=TRUE,full.names=FALSE)
rawRingDatas
```

```{r data cleaning}
for (d in 1:length(rawRingDatas)) {
  df_name <- rawRingDatas[d]
  title <- paste0(substr(df_name,1,2),"_",substr(df_name,4,7))
  
  treeinfofile <- list.files(here(wl),pattern=paste0(title,"_treeinfobai",".*\\.csv$"),full.names=FALSE)[1]
  
  
  rawRingData <- read.csv(here(wd,df_name))
  treeinfo <- read.csv(here(wl,treeinfofile)) %>%
    select(Tag,Lat,Long)
  
  rawRingData_long <- rawRingData %>%
  select(-X)%>%
  pivot_longer(
    cols = starts_with("X"),
    names_to = "years",
    names_prefix = "X",
    values_to = "Ring_width(mm)") %>%
  select(-Species,-Elev) %>%
    left_join(treeinfo,by="Tag")%>%
    mutate(species = substr(Tag,1,4))

  head(rawRingData_long)
  write.csv(rawRingData_long,here(wo,df_name))
}
```


Alberta data
```{r directory} 
alwd <- "analysis/pnwandmore/data/other_treerings/direct_searching/Alberta/dataCite/RawtreeRing/whiteSpruce1983-2016/treering_width.csv"
alwo <- 'analysis/pnwandmore/input/other_tree_rings/direct_searching/Alberta'
```

```{r data input}
alrrd <- read.csv(here(alwd))
```

```{r cleaning and saving}
alrrd <- alrrd %>%
  select(id,year,trw,Latitude,Longitude)%>%
  rename("Ring_width(mm)" = trw,
         Lat = Latitude,
         Long = Longitude,
         Tag = id,
         years = year)

alrrd$speciesfull <- "Picea glauca"
alrrd$species <- "PIGL"

write.csv(alrrd,here(alwo,"treering_width.csv"))
```

Arizona datasearch
```{r directory}
arw <- "analysis/pnwandmore/data/other_treerings/direct_searching/Arizona/datasearch/rawTreeRing/bog_tree/bog_tree_ring_data.txt"
arwo <- 'analysis/pnwandmore/input/other_tree_rings/direct_searching/Arizona'
```

```{r data cleaning}
Lat <- tibble(Lat = 64.6408)
Long <- tibble(Long = 	-148.3334833)
arrrd <- read.table(here(arw),sep="\t",skip = 2,col.names = c("years","Tag","Tree location","Ring_width","Temperature (ºC, Growing Season: May, June, July, Aug)","Precipitation (mm, Growing Season: May, June, July, Aug)")) %>%
  select(years,Tag,Ring_width)%>%
  rename("Ring_width(mm)"=Ring_width)%>%
  bind_cols(Lat,Long)

arrrd$speciesfull <- "Picea mariana"
arrrd$species <- "PIMA"

write.csv(arrrd,here(arwo,"bog_tree_ring_data.csv"))
```

```{r}
arw1 <- "analysis/pnwandmore/data/other_treerings/direct_searching/Arizona/datasearch/rawTreeRing/PIPOcores518/PIPOCores518RingWidths.txt"
arwlgeo <- "analysis/pnwandmore/data/other_treerings/direct_searching/Arizona/datasearch/rawTreeRing/PIPOcores518/PIPOCores518Meta.txt"

geoloc <- read_tsv(here(arwlgeo)) %>%
  select(CORE_CN,LON,LAT)
arrrd1 <- read.table(here(arw1),sep="\t",skip=1,
                     col.names = c("Ring_width","years","CORE_CN")) %>%
  left_join(geoloc,by = "CORE_CN")%>%
  rename(Lat = LAT,
         Long = LON,
         "Ring_width(mm)"=Ring_width,
         Tag = CORE_CN)
arrrd1$species <- "PIPO"
arrrd1$speciesfull <- "Pinus ponderosa"
write.csv(arrrd1,here(arwo,"PIPOCores518RingWidths.csv"))
```

Arizona dataCite
```{r}
arc <- "analysis/pnwandmore/data/other_treerings/direct_searching/Arizona/dataCite/RawTreeRing/speciesResponse/core_data_22dec2023 (1)/core_data_22dec2023 (1).csv"
arc1 <- read.csv(here(arc)) %>%
  select(Year,Series,RW,x,y)%>%
  rename(years = Year,
         Tag = Series,
         "Ring_width(mm)"=RW,
         Long = x,
         Lat = y)
arc1$species <- "PIPO"
arc1$speciesfull <- "Pinus ponderosa"
write.csv(arc1,here(arwo,"core_data_22dec2023.csv"))
```

BC datasearch
```{r} 
wdb <- "analysis/pnwandmore/data/other_treerings/direct_searching/BC/datasearch/RawTreeRing/EN22_Tree_Ring_Widths.tab"
savebc<-'analysis/pnwandmore/input/other_tree_rings/direct_searching/Bc'

species_key <- data.frame(
  speciesfull = c(
    "Picea mariana",
    "Larix laricina",
    "Betula papyrifera",
    "Picea glauca",
    "Picea sitchensis",
    "Populus tremuloides",
    "Pinus contorta",
    "Populus balsamifera",
    "Abies lasiocarpa",
    "Pinus ponderosa"
  ),
  species = c(
    "PIMA",
    "LALA",
    "BEPA",
    "PIGL",
    "PISI",
    "POTR",
    "PICO",
    "POBA",
    "ABLA",
    "PIPO"
  ),
  stringsAsFactors = FALSE
)


bcdata <- read_tsv(here(wdb),skip = 52) %>%
  rename(Lat=Latitude,
         Long=Longitude,
         "Ring_width(mm)"="TRW [mm]" ,
         years="Age [a AD/CE]",
         Tag="Subsample ID",
         speciesfull=Species)%>%
  select(Tag,Lat,Long,years,"Ring_width(mm)",speciesfull)%>%
  left_join(species_key,by="speciesfull")
write.csv(bcdata,here(savebc,"EN22_Tree_Ring_Widths.csv"))
#unique(bcdata$speciesfull)
#head(bcdata)
```

Colorado datasearch
*** Duke, combine,Bonaza left
```{r} 
#TreeBase2021
wdco <- "analysis/pnwandmore/data/other_treerings/direct_searching/colorado/datasearch/rawTreeRing/TreeBase2021/731f7ef6-1930-453c-b838-f06ec61431ac/data-EP-WC-RING-WIDTH-INDEX-Clemons.csv"

saveco<-"analysis/pnwandmore/input/other_tree_rings/direct_searching/Colorado/treeBase2021"


#cospatial <- read_csv(here(spatialco))
#cogrowth <- read_csv(here(growthco))
codata <- read_csv(here(wdco))%>%
  pivot_longer(cols=-1,
               names_to = "Tag",
               values_to = "Ring_width(mm)")%>%
  bind_cols(Long = -105.47203,
            Lat = 40.44473) %>%
  mutate(
    species = case_when(
      str_detect(Tag, regex("Pip", ignore_case = TRUE)) ~ "PIPO",
      str_detect(Tag, regex("Pien|Pie", ignore_case = TRUE)) ~ "PIEN",
      str_detect(Tag, regex("Psm", ignore_case = TRUE)) ~ "PSME",
      str_detect(Tag, regex("Unkn", ignore_case = TRUE)) ~ "UNKN",
      TRUE ~ NA_character_
    ),
    speciesfull = case_when(
      species == "PIPO" ~ "Pinus ponderosa",
      species == "PIEN" ~ "Picea engelmannii",
      species == "PSME" ~ "Pseudotsuga menziesii",
      species == "UNKN" ~ "Unkown",
      TRUE ~ NA_character_
    )
  )%>%
  rename(years = Year)
  
  #left_join(cogrowth,by="TREE_ID")%>%
  #left_join(cospatial,by=c("STATE","PROV"))%>%
  #select(YEAR,TREE_ID,"Ring_width(mm)",LAT,LONG)%>%
  #rename(years=YEAR,
         #Tag=TREE_ID,
         #Lat=LAT,
         #Long=LONG)

write.csv(codata,here(saveco,"ata-EP-WC-RING-WIDTH-INDEX-Clemons.csv"))
head(codata)
```
```{r}
#Duke
wdco1 <- "analysis/pnwandmore/data/other_treerings/direct_searching/colorado/datasearch/rawTreeRing/Dukes/edi.1969.1/Dukes_tree_rings.csv"

saveco1 <- "analysis/pnwandmore/input/other_tree_rings/direct_searching/Colorado/Duke"

codata1 <- read_csv(here(wdco1))%>%
  mutate(Tag = paste0(stem,side),
         species = toupper(sp),speciesfull = case_when(
      species == "PIGL" ~ "Picea glauca",
      species == "ACRU" ~ "Acer rubrum",
      species == "ACSA" ~ "Acer saccharum",
      species == "BEAL" ~ "Betula alleghaniensis",
      species == "FAGR" ~ "Fagus grandifolia",
      species == "OSVI" ~ "Ostrya virginiana",
      species == "TSCA" ~ "Tsuga canadensis",
      TRUE ~ NA_character_
    ))%>%
  select(ring.width,year,species,speciesfull,Tag)%>%
  rename("Ring_width(mm)"="ring.width",
         years=year)%>%
  bind_cols(Lat = 46.370,
            Long = -87.151)
head(codata1)
write.csv(codata1,here(saveco1,"Dukes_tree_rings.csv"))
```

```{r}
#Bonaza
wdco2 <- "analysis/pnwandmore/input/other_tree_rings/direct_searching/Colorado/BonanzaCreek2018/775_birch.csv"
codata2 <- read_csv(here(wdco2))%>%
  bind_cols(species = "BEPA",
            speciesfull = "Betula papyrifera")

wdco3 <- "analysis/pnwandmore/input/other_tree_rings/direct_searching/Colorado/BonanzaCreek2018/775_spruce.csv"
codata3 <- read_csv(here(wdco3))%>%
  bind_cols(species = "PIGL", speciesfull = "Picea glauca")

write.csv(codata2,here("analysis/pnwandmore/input/other_tree_rings/direct_searching/Colorado/BonanzaCreek2018","775_birch.csv"))
write.csv(codata3,here("analysis/pnwandmore/input/other_tree_rings/direct_searching/Colorado/BonanzaCreek2018","775_spruce.csv"))
```

```{r}
#combine
# no coordinates found
```

idaho dataCite
```{r}
wdid <- "analysis/pnwandmore/data/other_treerings/direct_searching/idaho/dataCite/rawTreeRing/coredata2023/core_data_22dec2023.csv"
saveid <- "analysis/pnwandmore/input/other_tree_rings/direct_searching/idaho/coredata2023"

iddata <- read_csv(here(wdid))%>%
  select(Series,Year,RW,x,y)%>%
  rename(Tag=Series,
         years=Year,
         "Ring_width(mm)"=RW,
         Long=x,
         Lat=y)%>%
  bind_cols()
write.csv(iddata,here(saveid,"core_data_22dec2023.csv"))
```
montana datasearch-coweetaLab
```{r}
wdmon <- "analysis/pnwandmore/data/other_treerings/direct_searching/montana/datasearch/rawTreeRing/CoweetaLab/Data"
savemon <- "analysis/pnwandmore/input/other_tree_rings/direct_searching/montana/coweetaLab"

montdatas <- list.files(here(wdmon),recursive=TRUE,full.names=FALSE)
for (i in 1:length(montdatas)) {
  dfname <- montdatas[i]
  mondata <- read.csv(here(wdmon,dfname),na.strings=".")%>%
  pivot_longer(cols=-1,
               names_to = "Tag",
               values_to = "Ring_width")%>%
  mutate("Ring_width(mm)"=Ring_width/1000)%>%
  rename(years=X)%>%
  select(-Ring_width)%>%
  bind_cols(LongW=-83.47845,LongE=-83.42166,LatN=35.07382,LatS=35.02734,species = substr(dfname,1,4))%>%
    mutate(speciesfull = case_when(
      species == "ACRU" ~ "Acer rubrum",
      species == "BELE" ~ "Betula lenta",
      species == "LITU" ~ "Liriodendron tulipifera",
      species == "QUAL" ~ "Quercus alba",
      species == "QUMO" ~ "Quercus montana",
      species == "QURU" ~ "Quercus rubra",
      TRUE ~ NA_character_
    ))
    
  write.csv(mondata,here(savemon,dfname))
}
```
montana datasearch-treeGrowthData
```{r}
wdmon1 <- "analysis/pnwandmore/data/other_treerings/direct_searching/montana/datasearch/rawTreeRing/Tree growth data/Treegrowthdata_/LSWL_2017_AVE_CNT_BAI.csv"
savemon1 <- "analysis/pnwandmore/input/other_tree_rings/direct_searching/montana/treeGrowth"
mondata1 <- read_csv(here(wdmon1))%>%
  rename(years=Year,
         Tag="Tree ID",
         "Ring_width(mm)"="Ave. Ringwidth")%>%
  select(years,Tag,"Ring_width(mm)")%>%
  bind_cols(LongW=-113.1207,
            LongE=-112.9285,
            LatN=44.9181,
            LatS=44.7155)
head(mondata1)
write.csv(mondata1,here(savemon1,"LSWL_2017_AVE_CNT_BAI.csv"))
```
montana dataCite
```{r}
wdmon2 <- "analysis/pnwandmore/data/other_treerings/direct_searching/montana/dataCite/rawTreeRing/coredata2023/core_data_22dec2023.csv"

savemon2 <- "analysis/pnwandmore/input/other_tree_rings/direct_searching/montana/core2023"

mondata2 <- read_csv(here(wdmon2))%>%
  rename(Long=x,
         Lat=y,
         "Ring_width(mm)"=RW,
         years=Year,
         Tag=Series)%>%
  select(years,Long,Lat,"Ring_width(mm)",Tag)
write.csv(mondata2,here(savemon2,"core_data_22dec2023.csv"))
head(mondata2)
```
nevada datasearch
"D:\ubc_study\udergrad_research\temporal ecology lab\climategrowthshifts\analysis\pnwandmore\data\other_treerings\direct_searching\nevada\datasearch\rawTreeRing\North_South_California_Tree_Ring_Study\northerncaliforniaringwidthdata" Don't know how to process RWL

nevada dataCite
```{r}
wdne <- "analysis/pnwandmore/data/other_treerings/direct_searching/nevada/dataCite/rawTreeRing/Northern Nevada aspen (Populus tremuloides) data"
savene <- "analysis/pnwandmore/input/other_tree_rings/direct_searching/nevada/Northern(noCoor)"

nedatas <- list.files(path=here(wdne),pattern="*ring*",full.names = FALSE)
for (i in 1:length(nedatas)) {
  df<- nedatas[i]
  nedata <- read_csv(here(wdne,df))%>%
  pivot_longer(
    cols= -c(1,2),
    names_to = "Tag",
    values_to="Ring_width(mm)"
  )%>%
  select(-"Mountain Range")%>%
  rename(years=Year)
  
  write.csv(nedata,here(savene,df))
}
```

New Maxico dataCite
```{r}
wdmx <- "analysis/pnwandmore/data/other_treerings/direct_searching/NewMexico/dataCite/rawTreeRing/doi_10_5061_dryad_fv322__v20180406/sample+tree+growth+data+with+locations.csv"

savemx <- "analysis/pnwandmore/input/other_tree_rings/direct_searching/NewMexico"

mxdata <- read_csv(here(wdmx))%>%
  select(Year,"Sample tree ID","Longitude","Latitude","Ring width (mm)",Species)%>%
  rename(years=Year,
         Tag="Sample tree ID",
         Long = Longitude,
         Lat = Latitude,
         "Ring_width(mm)"="Ring width (mm)",
         speciesfull = Species)%>%
  mutate(species = case_when(
      speciesfull == "Picea engelmannii" ~ "PIEN",
      speciesfull == "Abies lasiocarpa" ~ "ABLA",
      speciesfull == "Pinus contorta var. latifolia" ~ "PICO",
      speciesfull == "Pinus ponderosa" ~ "PIPO",
      speciesfull == "Pseudotsuga menziesii" ~ "PSME",
      TRUE ~ NA_character_
    ))
write.csv(mxdata,here(savemx,"sample+tree+growth+data+with+locations.csv"))
head(mxdata)
unique(mxdata$speciesfull)
```

Utah dataCite
```{r}
wdu <- "analysis/pnwandmore/data/other_treerings/direct_searching/Utah/datasearch/rawTreeRing/1/UT_rw.csv"
saveu <- "analysis/pnwandmore/input/other_tree_rings/direct_searching/Utah"

spcd_lookup <- tibble::tribble(
  ~SPCD, ~species, ~speciesfull,
  15,   "ABCO", "Abies concolor",
  19,   "ABLA", "Abies lasiocarpa",
  65,   "JUSC", "Juniperus scopulorum",
  93,   "PIEN", "Picea engelmannii",
  96,   "PISI", "Picea sitchensis",
  106,  "PIED", "Pinus edulis",
  108,  "PICO", "Pinus contorta",
  122,  "PIPO", "Pinus ponderosa",
  133,  "PIMO", "Pinus monophylla",
  202,  "PSME", "Pseudotsuga menziesii",
  321,  "ACGL", "Acer glabrum",
  NA,NA,NA
)


udata <- read_csv(here(wdu)) %>%
  mutate(Tag=paste(PLT_CN,TRE_CN,sep="-"))%>%
  rename(years=Year,
         "Ring_width(mm)"=RW) %>%
  select(years,Tag,"Ring_width(mm)",SPCD)%>%
  left_join(spcd_lookup,by = "SPCD")

write.csv(udata,here(saveu,"UT_rw.csv"))
head(udata)
#unique(udata$SPCD)
```
Wyoming datasearch 
"I don't know how to read this file Cottonwood_Wind_Ring_Widths"
