---
title: |
   | Tree growth in a 
   | non-stationnary climate
author: 'Project update---and moving forward!'
format:
  revealjs: 
    theme: default
    include-in-header:
          - text: |
              <style>
                #title-slide .title {
                  font-size: 60px;
                }
              </style>
---

## Current model

For $n$ observations for one tree $k$ of the species $sp$ on the stand $p$:

$$
\vec{\text{rw}}_{sp,p,k} = 
\begin{bmatrix}
  \text{rw}_{sp,p,k}(t_1) \\
  \text{rw}_{sp,p,k}(t_2) \\
  \vdots \\
  \text{rw}_{sp,p,k}(t_n)
\end{bmatrix}
$$ 

## Current model

For $n$ observations for one tree $k$ of the species $sp$ on the stand $p$:

$$\log\vec{\text{rw}}_{sp,p,k} \sim \text{multinormal}(\vec{\mu}_{sp,p} + \vec{f}_{\text{long},k}, \hspace{0.08cm} \vec{\Sigma})$$
 
## Current model

For $n$ observations for one tree $k$ of the species $sp$ on the stand $p$:

$$\log\vec{\text{rw}}_{sp,p,k} \sim \text{multinormal}(\vec{\mu}_{sp,p} + \color{#84AE92}{\vec{f}_{\text{long},k}}, \hspace{0.08cm} \vec{\Sigma})$$

$$\color{#84AE92}{\vec{f}_{\text{long},k}} \sim \text{multinormal}(\vec{0}, \vec{\mathrm{K}}_{sp})$$

## Long-term Gaussian process

One long-term scale Gaussian process $\color{#84AE92}{\vec{f}_{\text{long},k}}$ for each tree $k$,
to capture the long-term variation of tree growth, likely related to allometric constraints  

- one pair of parameters $\{\gamma_{\text{sp}}, \rho_{\text{sp}}\}$ per species $sp$, which defined one covariance matrix $\vec{\mathrm{K}}_{sp}$ per species
- prior mass of $\rho_{\text{sp}}$ between 20 and 60 years

## Current model

For $n$ observations for one tree $k$ of the species $sp$ on the stand $p$:

$$\log\vec{\text{rw}}_{sp,p,k} \sim \text{multinormal}(\vec{\mu}_{sp,p} + \vec{f}_{\text{long},k}, \hspace{0.08cm} \vec{\Sigma})$$

$$\vec{f}_{\text{long},k} \sim \text{multinormal}(\vec{0}, \vec{\mathrm{K}}_{sp})$$

## The climate behavior

::: {style="font-size: 70%;"}
$$
\vec{\mu}_{sp,p} = 
\begin{bmatrix}
  \mu_{sp,p}(t_1) \\
  \mu_{sp,p}(t_2) \\
  \vdots \\
  \mu_{sp,p}(t_n)
\end{bmatrix}
$$
:::

The elements $\mu_{sp,p}(t)$ are built up from [the short-term scale GP]{style="color:#FADA7A;"} and from contributions of [three climate variables]{style="color:#578FCA;"}:

::: {style="font-size: 70%;"}
$$ \begin{split} 
\forall i  \in [\![1,\mathrm{n}]\!]  \hspace{1cm} \mu_{sp,p}(t_i)= \alpha
    +  \color{#578FCA}{\beta_{\text{GDD}_{sp}} * \text{GDD}(p,t_i)}
    + \color{#578FCA}{\beta_{\text{SM}_{sp}}  * \text{SM}(p,t_i)}\\
    + \color{#578FCA}{\beta_{\text{VPD}_{sp}}  * \text{VPD}(p,t_i)}
    + \kappa_{\text{sp}}
    * \color{#FADA7A}{f_{\text{short,p}}(t_i)}
     \end{split} $$

$$\color{#FADA7A}{\vec{f}_{\text{short},p}} \sim \text{multinormal}(\vec{0}, \vec{\mathrm{K}}_\text{short})$$
:::

## Short-term Gaussian process

One short-term scale Gaussian process $\color{#FADA7A}{\vec{f}_{\text{short},p}}$ per stand $p$, to capture climatic variations that span several years

- two parameters $\{\gamma_{\text{short}}, \rho_{\text{short}}\}$, which defined one covariance matrix $\vec{\mathrm{K}}_\text{short}$ common across all stands
- prior mass of $\rho_{\text{short}}$ between 3 and 10 years
- does not depend on species, but its effect is scaled by a factor $\kappa_\text{sp}$


## Our climate predictors

Three climate variables, occuring at the annual-scale

- the growing degree days $\color{#578FCA}{\text{GDD}(p,t_i)}$ accumulated during the entire year, in $\text{x}100$ $^{\circ}\text{C}$
- the mean soil moisture $\color{#578FCA}{\text{SM}(p,t_i)}$ (weighted average across soil horizons) during summer, in $\%$
- the mean vapor pressure deficit $\color{#578FCA}{\text{VPD}(p,t_i)}$ during summer, in $\text{hPa}$

The slopes are species-specific, without any partial pooling (for now).

Variables were obtained and processed from the WLDAS dataset (1979–present).

## Tree-ring data
::: {style="font-size: 70%;"}
Tree-ring data were extracted from the ITRDB. 
We temporarily restricted our analysis to complete time series from 1980 to 2010, and to four species.
:::
```{r}
#| echo: false

wd <- "~/projects/climategrowthshifts/analysis/pnw"

library(rnaturalearth)
library(terra)
library(tidyterra)
library(ggplot2)
library(dplyr)

set.seed(5838299)

datasets <- readRDS(file = file.path(wd, 'data/itrdb', paste0('itrdb_info.rds')))
datasets <- datasets[datasets$last_year >= 2010,]
datasets <- datasets[datasets$dataset != 'wa_149',] # temporary
datasets <- datasets[datasets$species_code %in% c("PIPO", "TSME", "PSME", "THPL"),]

world <- vect(ne_countries(scale = "medium", returnclass = "sf"))
ggplot() +
  geom_spatvector(data = world, color = 'white', fill = 'grey90') +
  geom_point(data = datasets, aes(x = east_lon, y = north_lat), color = 'white', size = 2.5) +
  geom_point(data = datasets, aes(x = east_lon, y = north_lat, color = species_code)) +
  theme_minimal() +
  theme(axis.title = element_blank(), legend.title = element_blank()) +
  coord_sf(xlim = c(-127, -110), ylim = c(34,51)) +
  labs(title = paste0(length(unique(datasets$species_code)), ' species, ', length(unique(datasets$dataset)), ' sites'))

raw_data <- data.frame()
for(d in 1:nrow(datasets)){
  
  raw_data_d <- read.csv(file.path(wd, 'data/itrdb', datasets[d, 'state'], 'total_ring_width/data', datasets[d, 'dataset'], 'cleaned_tree_data_with_flags.csv'))
  
  coltokeep <-  !grepl(colnames(raw_data_d), pattern = 'outlier')
  raw_data_d <- raw_data_d[, coltokeep]
  
  raw_data_d <- raw_data_d[raw_data_d$age_CE >= 1980 & raw_data_d$age_CE <= 2010 & !is.na(raw_data_d$age_CE), ]
  raw_data_d <- raw_data_d[ , colSums(is.na(raw_data_d)) < length(1980:2010)]
  
  raw_data_long <- raw_data_d %>%
    tidyr::pivot_longer(cols = -age_CE, names_to = 'core_id', values_to ='rw_core')
  names(raw_data_long)[1] <- 'year'
  raw_data_long$stand <- datasets[d, 'dataset']
  raw_data_long$species <- datasets[d, 'species_code']
  
  raw_data_long[raw_data_long$rw_core %in% c(-8), 'rw_core'] <- NA # dealing with weird values in or_105 dataset
  
  raw_data_long$tree_id <- stringr::str_split_i(raw_data_long$core_id, pattern = '_', i = 1)
  raw_data_long$tree_id <- paste0(raw_data_long$stand, "_", substr(raw_data_long$tree_id,1,nchar(raw_data_long$tree_id)-1))
  
  raw_data_d <- raw_data_long %>%
    group_by(stand, species, tree_id, year) %>%
    summarise(rw_core_ave = mean(rw_core, na.rm = TRUE), .groups = 'drop')
  
  trees_to_remove <- unique(raw_data_d[is.na(raw_data_d$rw_core_ave) | raw_data_d$rw_core_ave == 0 , 'tree_id'])
  raw_data_d <- raw_data_d[!(raw_data_d$tree_id %in% trees_to_remove$tree_id),]
  
  raw_data <- rbind(raw_data, raw_data_d)
}

group_dataset <- datasets %>%
  group_by(round(north_lat,2), round(south_lat,2), round(east_lon,2), round(west_lon,2)) %>%
  mutate(group_dataset = paste0('S',cur_group_id())) %>%
  ungroup() %>%
  dplyr::select(dataset, group_dataset)
raw_data <- merge(raw_data, group_dataset, by.x = 'stand', by.y = 'dataset')

# Load climate data
clim_pred <- readRDS(file.path(wd, "output", "climate",  "climpredictors_ext_pnw_june2025.rds"))
clim_pred$soilmoist_jja <- clim_pred$soilmoist_jja*100 # in percentage
clim_pred$gdd <- clim_pred$gdd/100 # in x100 degC
clim_pred$vpd_jja <- clim_pred$vpd_jja # in hPa?

# Sizes
uniq_tree_ids <- unique(raw_data$tree_id)
N_trees <- length(uniq_tree_ids)

uniq_stand_ids <- unique(raw_data$group_dataset)
N_stands <- length(unique(raw_data$group_dataset))

uniq_species_ids <- unique(raw_data$species) #  Important that JUOC is first here!
N_species <- length(uniq_species_ids)

# Common years
all_years <- NULL
for (tid in uniq_tree_ids) {
  years <- raw_data[raw_data$tree_id == tid,]$year
  if (is.null(all_years))
    all_years <- years
  else
    all_years <- intersect(years, all_years)
}

N_all_years <- length(all_years)

# Format data into ragged arrays
log_rw_obs <- c()
gdd_obs <- c()
sm_obs <- c()
vpd_obs <- c()
years <- c()
all_years_idxs <- c()
N_years <- c()
stand_idxs <- c()
species_idxs <- c()

idx <- 1
tree_start_idxs <- c()
tree_end_idxs <- c()

for (tid in uniq_tree_ids) {
  
  raw_data_tree <- raw_data[raw_data$tree_id == tid & raw_data$year %in% all_years,]
  
  years_tree <- raw_data_tree$year
  all_years_idxs_tree <- sapply(years_tree, function(y) which(all_years == y))
  N_years_tree <- length(years_tree)
  if(N_years_tree > 31){stop()}
  
  log_rw_obs_tree <- sapply(years_tree, 
                            function(y) 
                              log(raw_data_tree$rw_core_ave[raw_data_tree$year == y][1]))
  log_rw_obs <- c(log_rw_obs, log_rw_obs_tree)
  
  gdd_obs_tree <- sapply(years_tree, 
                         function(y) 
                           as.numeric(clim_pred$gdd[clim_pred$year == y & clim_pred$dataset == raw_data_tree$stand[1]][1]))
  gdd_obs <- c(gdd_obs, gdd_obs_tree)
  
  sm_obs_tree <- sapply(years_tree, 
                        function(y) 
                          as.numeric(clim_pred$soilmoist_jja[clim_pred$year == y & clim_pred$dataset == raw_data_tree$stand[1]][1]))
  sm_obs <- c(sm_obs, sm_obs_tree)
  
  vpd_obs_tree <- sapply(years_tree, 
                        function(y) 
                          as.numeric(clim_pred$vpd_jja[clim_pred$year == y & clim_pred$dataset == raw_data_tree$stand[1]][1]))
  vpd_obs <- c(vpd_obs, vpd_obs_tree)
  
  years <- c(years, years_tree)
  all_years_idxs <- c(all_years_idxs, all_years_idxs_tree)
  N_years <- c(N_years, N_years_tree)
  
  stand_tree <- which(uniq_stand_ids == raw_data_tree$group_dataset[1])
  # stand_tree <- which(uniq_stand_ids == raw_data_tree$region[1])
  stand_idxs <- c(stand_idxs, stand_tree)
  
  species_tree <- which(uniq_species_ids == raw_data_tree$species[1])
  species_idxs <- c(species_idxs, species_tree)
  
  tree_start_idxs <- c(tree_start_idxs, idx)
  idx <- idx + N_years_tree
  tree_end_idxs <- c(tree_end_idxs, idx - 1)
}

  
```
::: {style="font-size: 60%;"}
As a first crude approach, we considered that close sites (approx. <1km) have the same short-term scale GP.
:::

## Posterior quantification
```{r}
#| echo: false
N <- length(years)
data <- mget(c('N', 'N_all_years', 'N_trees',
               'log_rw_obs', 'gdd_obs', 'sm_obs', 'vpd_obs',
               'all_years', 'years', 'all_years_idxs', 'N_years',
               'stand_idxs', 'N_stands',
               'species_idxs', 'N_species',
               'tree_start_idxs', 'tree_end_idxs'))
# fit <- stan(file=file.path(wd, 'model/stan/model6_with3predictors_pnw_differentbetas.stan'),
#             data=data, seed=5838299, 
#             chains = 4, cores = 4,
#             warmup=1000, iter=2024, refresh=10)

fit <- readRDS("/home/victor/projects/climategrowthshifts/analysis/pnw/output/model/fit_21june2025_4species.rds")
setwd(file.path(wd, 'model'))
util <- new.env()
source('mcmc_analysis_tools_rstan.R', local=util)
source('mcmc_visualization_tools.R', local=util)
setwd(wd)
```
::: {style="font-size: 70%;"}
HMC does not appear to have major issues quantifying the posterior distribution.
:::
::: {style="font-size: 60%;"}
```{r}
#| echo: true
diagnostics <- util$extract_hmc_diagnostics(fit)
util$check_all_hmc_diagnostics(diagnostics)

samples <- util$extract_expectand_vals(fit)
base_samples <- util$filter_expectands(samples,
                                       c('alpha', 'beta_gdd', 'beta_sm', 'beta_vpd', 
                                         'rho', 'gamma',
                                         'f_tilde_sh',
                                         'rho_sh', 'gamma_sh',
                                         'kappa_sh_free',
                                         'sigma'),
                                       check_arrays=TRUE)
util$check_all_expectand_diagnostics(base_samples)
```
:::

## Retrodictive check

::: {style="font-size: 70%;"}
Let's take 4 random trees:
:::

```{r}
par(mfrow=c(2, 2))
for (t in  sample(1:N_trees, 4)) {
  idxs <- tree_start_idxs[t]:tree_end_idxs[t]
  rw_names <- sapply(idxs,
                     function(n) paste0('log_rw_pred[', n, ']'))
  util$plot_conn_pushforward_quantiles(samples, rw_names, data$years[idxs],
                                       xlab="Year", ylab="Log Ring Width Per mm", 
                                       display_ylim=c(-4, 4),
                                       main=paste0("Stand ", uniq_stand_ids[stand_idxs[t]], 
                                                   ", species ", uniq_species_ids[species_idxs[t]]))
  points(data$years[idxs], data$log_rw_obs[idxs], pch=16, cex=1.0, col="white")
  points(data$years[idxs], data$log_rw_obs[idxs], pch=16, cex=0.8, col="black")
  #abline(v=1974, lwd=2, lty=2, col="#DDDDDD")
}
```

## Retrodictive check

::: {style="font-size: 70%;"}
Let's take 4 other trees:
:::

```{r}

par(mfrow=c(2, 2))
for (t in  sample(1:N_trees, 4)) {
  idxs <- tree_start_idxs[t]:tree_end_idxs[t]
  rw_names <- sapply(idxs,
                     function(n) paste0('log_rw_pred[', n, ']'))
  util$plot_conn_pushforward_quantiles(samples, rw_names, data$years[idxs],
                                       xlab="Year", ylab="Log Ring Width Per mm", 
                                       display_ylim=c(-4, 4),
                                       main=paste0("Stand ", uniq_stand_ids[stand_idxs[t]], 
                                                   ", species ", uniq_species_ids[species_idxs[t]]))
  points(data$years[idxs], data$log_rw_obs[idxs], pch=16, cex=1.0, col="white")
  points(data$years[idxs], data$log_rw_obs[idxs], pch=16, cex=0.8, col="black")
  #abline(v=1974, lwd=2, lty=2, col="#DDDDDD")
}
```

## Retrodictive check

::: {style="font-size: 70%;"}
6 random trees in stand S18, period 1995-2010:
:::

```{r}
trees_s4 <- which(stand_idxs == which(uniq_stand_ids == 'S18'))

par(mfrow=c(2, 3))
for (t in  sample(trees_s4, 6)) {
  idxs <- tree_start_idxs[t]:tree_end_idxs[t]
  idxs <- idxs[16:31]
  rw_names <- sapply(idxs,
                     function(n) paste0('log_rw_pred[', n, ']'))
  util$plot_conn_pushforward_quantiles(samples, rw_names, data$years[idxs],
                                       xlab="Year", ylab="Log Ring Width Per mm", 
                                       display_ylim=c(-4, 4),
                                       main=paste0("Stand ", uniq_stand_ids[stand_idxs[t]], 
                                                   ", species ", uniq_species_ids[species_idxs[t]]))
  points(data$years[idxs], data$log_rw_obs[idxs], pch=16, cex=1.0, col="white")
  points(data$years[idxs], data$log_rw_obs[idxs], pch=16, cex=0.8, col="black")
  #abline(v=1974, lwd=2, lty=2, col="#DDDDDD")
}
```

## Retrodictive check

::: {style="font-size: 70%;"}
What about our climate predictors?
:::

```{r}
#| eval: false

par(mfrow=c(2, 2), mar = c(4,4.5,2.5,1))
for(spind in 1:4){
  start <- data$tree_start_idxs[min(which(data$species_idxs == spind))]
  end <- data$tree_end_idxs[max(which(data$species_idxs == spind))]
  pred_names <- sapply(start:end, function(n) paste0('log_rw_pred[', n, ']'))
  util$plot_conditional_median_quantiles(samples, pred_names, data$gdd_obs[start:end],
                                       0, 32, 1, data$log_rw_obs[start:end], 
                                       residual=FALSE,
                                       ylab = 'Marginal quantiles',
                                       xlab="GDD (x100degC)", main = paste0('Species ', uniq_species_ids[spind]))
}

```

## Retrodictive check

::: {style="font-size: 70%;"}
What about our climate predictors?
:::

```{r}
#| eval: false

par(mfrow=c(2, 2), mar = c(4,4.5,2.5,1))
for(spind in 1:4){
  start <- data$tree_start_idxs[min(which(data$species_idxs == spind))]
  end <- data$tree_end_idxs[max(which(data$species_idxs == spind))]
  pred_names <- sapply(start:end, function(n) paste0('log_rw_pred[', n, ']'))
  util$plot_conditional_median_quantiles(samples, pred_names, data$sm_obs[start:end],
                                       10, 34, 1, data$log_rw_obs[start:end], 
                                       residual=FALSE,
                                       ylab = 'Marginal quantiles',
                                       xlab="SM (%)", main = paste0('Species ', uniq_species_ids[spind]))
}

```

## Retrodictive check

::: {style="font-size: 70%;"}
What about our climate predictors?
:::

```{r}
#| eval: false

par(mfrow=c(2, 2), mar = c(4,4.5,2.5,1))
for(spind in 1:4){
  start <- data$tree_start_idxs[min(which(data$species_idxs == spind))]
  end <- data$tree_end_idxs[max(which(data$species_idxs == spind))]
  pred_names <- sapply(start:end, function(n) paste0('log_rw_pred[', n, ']'))
  util$plot_conditional_median_quantiles(samples, pred_names, data$vpd_obs[start:end],
                                       0, 22, 1, data$log_rw_obs[start:end], 
                                       residual=FALSE,
                                       ylab = 'Marginal quantiles',
                                       xlab="VPD (hPa)", main = paste0('Species ', uniq_species_ids[spind]))
}

```


## Inference
```{r}
par(mfrow=c(2, 2), mar = c(4,4,1,1))
for (s in 1:data$N_species) {
  util$plot_expectand_pushforward(samples[[ paste0('rho[', s, ']')]], 250,
                                  flim = c(0,50),
                                  display_name= paste0("rho ", uniq_species_ids[s]))
  xs <- seq(0, 60, 0.01)
  ys <- dlnorm(xs, 3.55, 0.24)
  lines(xs, ys, lwd=2, col=util$c_light)
}
```

## Inference
```{r}
par(mfrow=c(1, 1), mar = c(4,4.5,1,1))
names <- sapply(1:data$N_species,
                function(sp) paste0('beta_gdd[', sp, ']'))
util$plot_disc_pushforward_quantiles(samples, names,
                                     xlab="Species",
                                     xticklab=uniq_species_ids,
                                     ylab=expression(beta[GDD]))
```

## Inference
```{r}
par(mfrow=c(1, 1), mar = c(4,4.5,1,1))
names <- sapply(1:data$N_species,
                function(sp) paste0('beta_sm[', sp, ']'))
util$plot_disc_pushforward_quantiles(samples, names,
                                     xlab="Species",
                                     xticklab=uniq_species_ids,
                                     ylab=expression(beta[SMoist]))
```

## Inference
```{r}
par(mfrow=c(1, 1), mar = c(4,4.5,1,1))
names <- sapply(1:data$N_species,
                function(sp) paste0('beta_vpd[', sp, ']'))
util$plot_disc_pushforward_quantiles(samples, names,
                                     xlab="Species",
                                     xticklab=uniq_species_ids,
                                     ylab=expression(beta[VPD]))
```

## Pair plots

```{r}
par(mfrow=c(2, 2), mar = c(4,4.5,2.5,1))
s <- 1
plot(c(samples[[paste0("beta_gdd[",s,"]")]]) ~ c(samples[[paste0("beta_vpd[",s,"]")]]),
     xlab = expression(beta[VPD]), ylab = expression(beta[GDD]), 
     main = paste0('Species ', uniq_species_ids[1]))
s <- 2
plot(c(samples[[paste0("beta_gdd[",s,"]")]]) ~ c(samples[[paste0("beta_vpd[",s,"]")]]),
     xlab = expression(beta[VPD]), ylab = expression(beta[GDD]), 
     main = paste0('Species ', uniq_species_ids[s]))
s <- 3
plot(c(samples[[paste0("beta_gdd[",s,"]")]]) ~ c(samples[[paste0("beta_vpd[",s,"]")]]),
     xlab = expression(beta[VPD]), ylab = expression(beta[GDD]), 
     main = paste0('Species ', uniq_species_ids[s]))
s <- 4
plot(c(samples[[paste0("beta_gdd[",s,"]")]]) ~ c(samples[[paste0("beta_vpd[",s,"]")]]),
     xlab = expression(beta[VPD]), ylab = expression(beta[GDD]), 
     main = paste0('Species ', uniq_species_ids[s]))
```   

## Pair plots

```{r}
par(mfrow=c(2, 2), mar = c(4,4.5,2.5,1))
s <- 1
plot(c(samples[[paste0("beta_gdd[",s,"]")]]) ~ c(samples[[paste0("beta_sm[",s,"]")]]),
     xlab = expression(beta[SMoist]), ylab = expression(beta[GDD]), 
     main = paste0('Species ', uniq_species_ids[1]))
s <- 2
plot(c(samples[[paste0("beta_gdd[",s,"]")]]) ~ c(samples[[paste0("beta_sm[",s,"]")]]),
     xlab = expression(beta[SMoist]), ylab = expression(beta[GDD]), 
     main = paste0('Species ', uniq_species_ids[s]))
s <- 3
plot(c(samples[[paste0("beta_gdd[",s,"]")]]) ~ c(samples[[paste0("beta_sm[",s,"]")]]),
     xlab = expression(beta[SMoist]), ylab = expression(beta[GDD]), 
     main = paste0('Species ', uniq_species_ids[s]))
s <- 4
plot(c(samples[[paste0("beta_gdd[",s,"]")]]) ~ c(samples[[paste0("beta_sm[",s,"]")]]),
     xlab = expression(beta[SMoist]), ylab = expression(beta[GDD]), 
     main = paste0('Species ', uniq_species_ids[s]))
```   

## 

```{r}

par(mfrow=c(2, 2), mar = c(4,4.5,2.5,1))
for (spind in 1:data$N_species) {
  start <- data$tree_start_idxs[min(which(data$species_idxs == spind))]
  end <- data$tree_end_idxs[max(which(data$species_idxs == spind))]
  
  plot(data$gdd_obs[start:end] ~ data$vpd_obs[start:end], ylab = 'GDD', xlab = 'VPD', main = paste0('Species ', uniq_species_ids[spind]))
}


```  

## 

```{r}

par(mfrow=c(2, 4), mar = c(4,4.5,2.5,1))
data$tree_id <- rep(data$N_trees, each= length(data$all_years))
for (spind in 1:data$N_species) {
  start <- data$tree_start_idxs[min(which(data$species_idxs == spind))]
  end <- data$tree_end_idxs[max(which(data$species_idxs == spind))]
  gdd <- sapply(start:end, function(i){
    median(samples[[paste0("beta_gdd[",spind,"]")]] * data$gdd_obs[i])
  })
  
  vpd <- sapply(start:end, function(i){
    median(samples[[paste0("beta_vpd[",spind,"]")]] * data$vpd_obs[i])
  })
  
  sum <- sapply(start:end, function(i){
    median(samples[[paste0("beta_gdd[",spind,"]")]] * data$gdd_obs[i] + samples[[paste0("beta_vpd[",spind,"]")]] * data$vpd_obs[i])
  })
  
  plot(gdd ~ vpd, ylab = expression(beta[GDD]%*%GDD), xlab = expression(beta[VPD]%*%VPD), main = paste0('Species ', uniq_species_ids[spind]))
  hist(sum, main ='', xlab = expression(beta[GDD]%*%GDD+beta[VPD]%*%VPD))
}


```   





