---
title: "Marginalizing shocks"
pdf-engine: lualatex
format:
  pdf:
    keep-tex: true
    fig-width: 7
    fig-height: 3
    code-overflow: wrap
monofont: 'Latin Modern Mono'
monofontoptions: 'Scale=0.75'
knitr:
  opts_chunk:
    comment: ''
    cache.lazy: FALSE
  opts_knit:
    global.par: TRUE
geometry:
  - top=20mm
  - left=20mm
  - bottom=20mm
  - right=20mm
---

*Cover that shocks which I ought not to see.* 

```{r}
#| echo: false
#| warning: false
#| message: false
wd <- "~/projects/climategrowthshifts/analysis/pnwandmore"
setwd(file.path(wd, 'model/shocks'))
util <- new.env()
source('mcmc_analysis_tools_rstan.R', local=util)
source('mcmc_visualization_tools.R', local=util)
# source('functions/custom_functions.R', local=util)
setwd(wd)
library(rstan)
options(mc.cores = parallel::detectCores())

chunk_hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- chunk_hook(x, options)
  paste0("\\linespread{0.5}\n", x, "\n\n\\linespread{2}")
})

par(family="serif", las=1, bty="l",
    cex.axis=0.8, cex.lab=0.85, cex.main=1,
    xaxs="i", yaxs="i", mar = c(5, 5, 3, 1))
```

## Model reparameterization

We marginalize the shocks in the mixture model!
Going from:
```{stan output.var = 'foo'}
#| eval: false
# Non-shocky
target += normal_lpdf(y[n] | mu[n] + delta_sck[n], sigma)
target += normal_lpdf(delta_sck[n] | 0, inner_tau_sck)

# Shocky
target += normal_lpdf(y[n] | mu[n] + delta_sck[n], sigma)
target += normal_lpdf(delta_sck[n] | 0, outer_tau_sck)
```

to a log-likelihood where shocks are marginalized out:
```{stan output.var = 'foo'}
#| eval: false
# Non-shocky
target += normal_lpdf(y[n] | mu[n], eta)

# Shocky
target += normal_lpdf( y[n] | mu[n], eta * sqrt{1 + nu / eta^2})
```

where:
$$\eta = \sqrt{ \sigma^2 + {\tau_\text{inner}^\text{shock}}^2 }$$
$$\nu = {\tau_\text{outer}^\text{shock}}^2-{\tau_\text{inner}^\text{shock}}^2$$
In Stan, we now have in the model block:

```{r}
writeLines(
  readLines(file.path(wd, 'model', 'stan', 
                      'model9_marginalshocks.stan'))[230:315])
```

## But... my shocks...

Our large shocks are likely tied to interesting ecological stuff... But by marginalizing them, we don't have inferences on them anymore. \
It could be doable to "*to marginalize out eta in the “small shock” branch of the mixture model but leave eta explicit in the “large shock” branch of the mixture model*" (Mike). \
Since I'm not totally sure how to do that, I decided to sample the $\delta_\text{shock}$ afterwards, to recover their posterior distributions in the generated quantities block. This is an ongoing project, so for now I recover only the shock state ($z_t$).

```{r}
writeLines(
  readLines(file.path(wd, 'model', 'stan', 
                      'model9_marginalshocks.stan'))[351:405])
```

## Test with a small dataset

Let's run this new model on a very small dataset, 21 trees and 1 site. \
The good thing here is that I don't need to tune the (non-)centered parametrization anymore.

```{r}
data <- readRDS(file = file.path(wd, 'output', 'model', 'data_15nov2025_az592.rds'))
data$N_stand_trees <- array(data$N_stand_trees, dim = 1)
data$N_stand_years <- array(data$N_stand_years, dim = 1)
data$stand_start_years_idxs <- array(data$stand_start_years_idxs, dim = 1)
data$stand_trees_end_idxs <- array(data$stand_trees_end_idxs, dim = 1)
data$stand_trees_start_idxs <- array(data$stand_trees_start_idxs, dim = 1)

if(FALSE){
  fit <- stan(file=file.path(wd, 'model', 'stan', 'model9_marginalshocks.stan'),
            data=data, seed=5838293, chains = 4,
            warmup=1000, iter=2024, refresh=10)
  saveRDS(fit, file.path(wd, 'model/shocks/output/marginal', 'fit.rds'))
}else{
  fit <- readRDS(file.path(wd, 'model/shocks/output/marginal', 'fit.rds'))
}

```

Let's look at the diagnostics:

```{r}
diagnostics <- util$extract_hmc_diagnostics(fit)
util$check_all_hmc_diagnostics(diagnostics)

samples <- util$extract_expectand_vals(fit)
base_samples <- util$filter_expectands(samples,
                                       c('alpha', 
                                         'beta_gdd', 'beta_sm', 'beta_vpd',
                                         'rho_sh', 'gamma_sh',
                                         'rho_sp', 'gamma_sp',
                                         'eta', 'nu',
                                         'phi_sck', 'omega_conc_sck', 'omega_nonconc_sck'),
                                       check_arrays = TRUE)
util$check_all_expectand_diagnostics(base_samples)

```

Clean! Let's continue. \
These are the latent shock states:

```{r}
#| fig-width: 7
#| fig-height: 4
set.seed(123456)
random_trees <- sample(1:data$N_trees, 4)

par(mfrow = c(2,2), mar = c(2, 5, 3, 1))
for(t in random_trees){
  idxs_t <- data$tree_start_idxs[t]:data$tree_end_idxs[t]
  names <- sapply(idxs_t,
                  function(x) paste0('sck_state[', x, ']'))
  util$plot_conn_pushforward_quantiles(samples, names, data$years[idxs_t],
                                       xlab="", ylab=paste0("Shock state - tree ",t), 
                                       display_ylim=c(-0.1, 1.1), display_xlim = c(1980, 2010))
}

```

## Next step

Is it possible to recover the posterior of $\delta_\text{shock}(k) ~|~ y(k)$, without having ${\tau_\text{inner}^\text{shock}}$ or ${\tau_\text{outer}^\text{shock}}$? \
Maybe a first step could be to fix: ${\tau_\text{inner}^\text{shock}} = 0$?

Because for now my predictions look like:

```{r}
#| fig-width: 7
#| fig-height: 4


par(mfrow = c(2,2), mar = c(2, 5, 3, 1))
for(t in random_trees){
  idxs_t <- data$tree_start_idxs[t]:data$tree_end_idxs[t]
  names <- sapply(idxs_t,
                  function(x) paste0('log_rw_pred[', x, ']'))
  util$plot_conn_pushforward_quantiles(samples, names, data$years[idxs_t],
                                       xlab="", ylab=paste0("Shock state - tree ",t), 
                                       display_ylim=c(-10, 5), display_xlim = c(1980, 2010))
  points(data$years[idxs_t], data$log_rw_obs[idxs_t], pch=16, cex=0.7, col="white")
  points(data$years[idxs_t], data$log_rw_obs[idxs_t], pch=16, cex=0.4, col="black")
}

```

## Assuming that tau_inner = 0

If I assume that ${\tau_\text{inner}^\text{shock}} = 0$, I can reconstruct the large shocks using the normal-normal conjugancy (*merci* Google):
$$\delta_\text{shock}(k) ~|~ y_k \sim \text{normal} \left( ({\tau_\text{outer}^\text{shock}}^2 / ({\tau_\text{outer}^\text{shock}}^2 + \sigma^2)) \times y_k ~,~ \sqrt{({\tau_\text{outer}^\text{shock}}^2 \times \sigma^2) / ({\tau_\text{outer}^\text{shock}}^2 + \sigma^2)}\right)$$

In the generated quantities block, we can then have:

```{r}
writeLines(
  readLines(file.path(wd, 'model', 'stan', 
                      'model9_marginalshocks_zerotauinner.stan'))[395:405])
```


```{r}

if(FALSE){
  fit2 <- stan(file=file.path(wd, 'model', 'stan', 'model9_marginalshocks_zerotauinner.stan'),
            data=data, seed=5838293, chains = 4,
            warmup=1000, iter=2024, refresh=10)
  saveRDS(fit2, file.path(wd, 'model/shocks/output/marginal', 'fit2.rds'))
}else{
  fit2 <- readRDS(file.path(wd, 'model/shocks/output/marginal', 'fit2.rds'))
}

```

Diagnostics are cleaned:
```{r}
diagnostics <- util$extract_hmc_diagnostics(fit2)
util$check_all_hmc_diagnostics(diagnostics)

samples <- util$extract_expectand_vals(fit2)
base_samples <- util$filter_expectands(samples,
                                       c('alpha', 
                                         'beta_gdd', 'beta_sm', 'beta_vpd',
                                         'rho_sh', 'gamma_sh',
                                         'rho_sp', 'gamma_sp',
                                         'sigma', 'outer_tau_sck',
                                         'phi_sck', 'omega_conc_sck', 'omega_nonconc_sck'),
                                       check_arrays = TRUE)
util$check_all_expectand_diagnostics(base_samples)

```

We can look at the latent shocks:
```{r}
#| fig-width: 7
#| fig-height: 4


par(mfrow = c(2,2), mar = c(2, 5, 3, 1))
for(t in random_trees){
  idxs_t <- data$tree_start_idxs[t]:data$tree_end_idxs[t]
  names <- sapply(idxs_t,
                  function(x) paste0('delta_sck[', x, ']'))
  util$plot_conn_pushforward_quantiles(samples, names, data$years[idxs_t],
                                       xlab="", ylab=paste0("Shock state - tree ",t), 
                                       display_ylim=c(-10, 0.1), display_xlim = c(1980, 2010))
}

```

And at the predictions:
```{r}
#| fig-width: 7
#| fig-height: 4


par(mfrow = c(2,2), mar = c(2, 5, 3, 1))
for(t in random_trees){
  idxs_t <- data$tree_start_idxs[t]:data$tree_end_idxs[t]
  names <- sapply(idxs_t,
                  function(x) paste0('log_rw_pred[', x, ']'))
  util$plot_conn_pushforward_quantiles(samples, names, data$years[idxs_t],
                                       xlab="", ylab=paste0("log(rw) - tree ",t), 
                                       display_ylim=c(-12, 1), display_xlim = c(1980, 2010))
   points(data$years[idxs_t], data$log_rw_obs[idxs_t], pch=16, cex=0.7, col="white")
  points(data$years[idxs_t], data$log_rw_obs[idxs_t], pch=16, cex=0.4, col="black")
}

```
\newpage
However, it seems that the shocks are bit too large...

```{r}
#| fig-width: 7
#| fig-height: 4


par(mfrow = c(2,2), mar = c(2, 5, 3, 1))
for(t in random_trees){
  idxs_t <- data$tree_start_idxs[t]:data$tree_end_idxs[t]
  names <- sapply(idxs_t,
                  function(x) paste0('log_rw_pred[', x, ']'))
  util$plot_conn_pushforward_quantiles(samples, names, data$years[idxs_t], 
                                       baseline_values = data$log_rw_obs[idxs_t],
                                       xlab="", ylab=paste0("Residuals - tree ",t), 
                                       display_ylim=c(-3, 1), display_xlim = c(1980, 2010),
                                       residual = TRUE)
}

```

And, let's not forget the stand-level GP, which is still wiggly:

```{r}
#| fig-width: 7
#| fig-height: 4

s <- 1
names <- sapply(1:data$N_all_years,
                  function(y) paste0('f_sh[', s, ',', y, ']'))
util$plot_realizations(samples, names, plot_xs = data$all_years[1:data$N_all_years], N_plots = 50,
                       main = paste0('Stand ', data$uniq_stand_ids[s]), ylab = 'f_sh',
                       display_xlim = data$all_years[c(1,data$N_all_years)], display_ylim = c(-2,2))
```