---
title: "Marginalizing shocks:\nA step back to stay on track" 
toc: true
number-sections: true
highlight: pygments
pdf-engine: lualatex
format:
  html:
    html-math-method: katex
    theme:
      - lux
      - custom.scss
    standalone: true
    embed-resources: true
    code-overflow: wrap
    linkcolor: "#B97C7C"
    fontsize: '11pt'
    code-block-border-left: "#8F2727"
    code-block-bg: "#f8f8f8"
    highlight-style: github
  pdf:
    keep-tex: true
    fig-width: 4
    fig-height: 4
    code-overflow: wrap
    monofont: 'Latin Modern Mono'
    monofontoptions:
      - Scale=0.5
format-links: false
knitr:
  opts_chunk:
    comment: ''
    cache.lazy: FALSE
  opts_knit:
    global.par: TRUE
geometry:
  - top=20mm
  - left=20mm
  - bottom=20mm
  - right=20mm
---

```{css, echo=FALSE}
.cell-output { border-left: #B97C7C 2px solid; border-radius: 0px;}
```

*Cover that shocks which I ought not to see.* 

```{r}
#| echo: false
#| warning: false
#| message: false
wd <- "~/projects/climategrowthshifts/analysis/pnwandmore"
setwd(file.path(wd, 'model/shocks'))
util <- new.env()
source('mcmc_analysis_tools_rstan.R', local=util)
source('mcmc_visualization_tools.R', local=util)
# source('functions/custom_functions.R', local=util)
setwd(wd)
library(rstan)
options(mc.cores = parallel::detectCores())

chunk_hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- chunk_hook(x, options)
  paste0("\\linespread{0.5}\n", x, "\n\n\\linespread{2}")
})

par(family="serif", las=1, bty="l",
    cex.axis=0.8, cex.lab=0.85, cex.main=1,
    xaxs="i", yaxs="i", mar = c(5, 5, 3, 1))

logit <- function(x) log(x/(1-x))
inv_logit <- function(x) return(exp(x) / (1 + exp(x)))
```

# Model reparameterization

We marginalize the shocks in the mixture model!
Going from:
```{stan output.var = 'foo'}
#| eval: false
# Non-shocky
target += normal_lpdf(y[n] | mu[n] + delta_sck[n], sigma)
target += normal_lpdf(delta_sck[n] | 0, inner_tau_sck)

# Shocky
target += normal_lpdf(y[n] | mu[n] + delta_sck[n], sigma)
target += normal_lpdf(delta_sck[n] | 0, outer_tau_sck)
```

to a log-likelihood where shocks are marginalized out:
```{stan output.var = 'foo'}
#| eval: false
# Non-shocky
target += normal_lpdf(y[n] | mu[n], eta)

# Shocky
target += normal_lpdf( y[n] | mu[n], eta * sqrt{1 + nu / eta^2})
```

where:
$$\eta = \sqrt{ \sigma^2 + {\tau_\text{inner}^\text{shock}}^2 }$$
$$\nu = {\tau_\text{outer}^\text{shock}}^2-{\tau_\text{inner}^\text{shock}}^2$$

:::{.callout-tip collapse=true title="Expand to see he new model block"}
```{r}
writeLines(
  readLines(file.path(wd, 'model', 'stan', 
                      'model9_marginalshocks.stan'))[230:315])
```
:::

# Recover the shocks afterwards

Our large shocks are likely tied to interesting ecological stuff... But by marginalizing them, we don't have inferences on them anymore. \
It's easy to recover the shock state, but what about the shock amplitude?\
If I assume that ${\tau_\text{inner}^\text{shock}} = 0$, I can reconstruct the large shocks using the normal-normal conjugancy:
$$\delta_\text{shock}(k) ~|~ y_k \sim \text{normal} \left( ({\tau_\text{outer}^\text{shock}}^2 / ({\tau_\text{outer}^\text{shock}}^2 + \sigma^2)) \times y_k ~,~ \sqrt{({\tau_\text{outer}^\text{shock}}^2 \times \sigma^2) / ({\tau_\text{outer}^\text{shock}}^2 + \sigma^2)}\right)$$

# Time to deal with those zeros!

Before digging too much on the shocks, I should first consider to deal with the missing rings.\
We have domain expertise to say that zeros are a signature of extreme tree stress and very low growth.
These missing rings correspond to observations that are below the measurement precision, which should be around $1\mathrm{e}{-3} ~\text{mm}$! We should thus consider that the probability of having a missing ring corresponds to the probability of getting values from `-Inf` to some threshold `epsilon` (which can be fixed, or a parameter to infer). The pseudo-code would be:

```{stan output.var = 'foo'}
#| eval: false

if(rw_obs[tree_idxs[y]] >= epsilon){
  target += normal_lpdf(log(rw_obs[tree_idxs[y]]) | mu[n], sqrt(outer_tau2_aux+sigma^2))
}else{
  target += normal_lcdf(log(epsilon)) | mu[n], sqrt(outer_tau2_aux+sigma^2))
}
```

:::{.callout-warning}
Here I used the condition `rw_obs[tree_idxs[y]] >= epsilon`, because `epsilon` is the "last" observed value before zero. I do have a minor concern, though: `log(epsilon)` is also included in the normal_lcdf integral... Is it a big issue?
:::

I modified the Stan code to include this censoring.

:::{.callout-tip collapse=true title="Expand to see a portion of the new model block"}
```{r}
writeLines(
  readLines(file.path(wd, 'model', 'stan', 
                      'model9_marginalshocks_zerotauinner_censoring_onlyconcordant.stan'))[310:326])
```
:::

But I am less sure on how to handle this in the `generated quantities` block. \
For now, when the observation $y$ is below the threshold $\epsilon$, I sample a $\log(y)$ in a truncated normal distribution (between $-\infty$ and $\log(\epsilon)$).

:::{.callout-tip collapse=true title="Expand to see a portion of the new generated quantities block"}
```{r}
writeLines(
  readLines(file.path(wd, 'model', 'stan', 
                      'model9_marginalshocks_zerotauinner_censoring_onlyconcordant.stan'))[394:411])
```
:::

I defined the truncated normal sampling function with the help of Stan manual.\
I add a condition to deal with `p_ub = 0`, when the probability that a sample from `normal(mu, sigma)` is below `ub` is zero. In this case, I consider that the entire density collapses to a single point, `ub` (i.e. $\log(\epsilon)$).

```{r}
writeLines(
  readLines(file.path(wd, 'model', 'stan', 
                      'model9_marginalshocks_zerotauinner_censoring_onlyconcordant.stan'))[57:67])
```

Currently I just set: `real epsilon = 1e-3;`. But we could fit this parameter later.

# A first fit on two stands

Here, we consider that:

- we have a normal population model for the shocks, with ${\tau_\text{inner}^\text{shock}} = 0$ (see above)  
- $\phi_\text{shock}$, the stand-level shock probability, varies across stands (heterogeneous)  
- $\omega_\text{shock}^\text{concordant}$, the tree-level shock probability given the stand is shocking, does not vary across stands (homogeneous)  
- $\omega_\text{shock}^\text{non-concordant} = 0$ (no tree-level shock without a stand-level shock)  

```{r}
data <- readRDS(file = file.path(wd, 'output', 'model', 'data_15nov2025_az592_az628.rds'))
data$rw_obs <- ifelse(data$log_rw_obs == log(1e-4), 0, exp(data$log_rw_obs))

if(FALSE){
  fit <- stan(file=file.path(wd, 'model', 'stan', 'model9_marginalshocks_zerotauinner_censoring.stan'),
            data=data, seed=5838293, chains = 4,
            warmup=1000, iter=2024, refresh=10)
  saveRDS(fit, file.path(wd, 'model/shocks/output/marginal', 'fit_normal_censoring.rds'))
}else{
  fit <- readRDS(file.path(wd, 'model/shocks/output/marginal', 'fit_normal_censoring.rds'))
}
``` 

Woaa! A flood of tree depth warnings.

```{r}
diagnostics <- util$extract_hmc_diagnostics(fit)
util$check_all_hmc_diagnostics(diagnostics)

samples <- util$extract_expectand_vals(fit)
base_samples <- util$filter_expectands(samples,
                                       c('alpha', 
                                         'beta_gdd', 'beta_sm', 'beta_vpd',
                                         'rho_sh', 'gamma_sh',
                                         'rho_sp', 'gamma_sp',
                                         'sigma', 
                                         'phi_sck', 'omega_conc_sck'),
                                       check_arrays = TRUE)
util$check_all_expectand_diagnostics(base_samples)
``` 

Let's do some retrodictive checks! Let's look at 4 trees on the first stand:

```{r}
#| fig-width: 8
#| fig-height: 9.4
#| echo: false

par(mfcol = c(4,2), cex.lab = 1.5)
for(t in 1:4){
  idxs_t <- data$tree_start_idxs[t]:data$tree_end_idxs[t]
  names <- sapply(idxs_t,
                  function(x) paste0('log_rw_pred[', x, ']'))
  util$plot_conn_pushforward_quantiles(samples, names, data$years[idxs_t],
                                       xlab="", ylab='Predictions (log)', main = paste0('Tree ', data$uniq_tree_ids[t]),
                                       display_ylim=c(-8, 2), display_xlim = c(1980, 2010))
  points(data$years[idxs_t], ifelse(data$log_rw_obs[idxs_t] == log(1e-4), NA, data$log_rw_obs[idxs_t]), pch=16, cex=1, col="white")
  points(data$years[idxs_t], ifelse(data$log_rw_obs[idxs_t] == log(1e-4), NA, data$log_rw_obs[idxs_t]), pch=16, cex=0.6, col="black")
  abline(v =  c(2002,1996), lty = 3, col="darkgrey")
  
  names <- sapply(idxs_t,
                  function(x) paste0('sck_state[', x, ']'))
  ticks <- c(unlist(lapply(seq(1980, 2000,10), function(x) c(x, rep(NA, 9)))), '2010')
  util$plot_disc_pushforward_quantiles(samples, names, data$years[idxs_t],
                                       xlab="", ylab='Shock state', 
                                       display_ylim=c(-0.1, 1.1), 
                                       xticklabs =  ticks)
  abline(v =  c(2002,1996), lty = 3, col="darkgrey")
  
  names <- sapply(idxs_t,
                  function(x) paste0('delta_sck[', x, ']'))
  util$plot_conn_pushforward_quantiles(samples, names, data$years[idxs_t],
                                       xlab="", ylab=expression(delta[shock]), 
                                       display_ylim=c(-8, 3), display_xlim = c(1980, 2010))
  abline(v =  c(2002,1996), lty = 3, col="darkgrey")
  
  names <- sapply(1:data$N_all_years,
                  function(x) paste0('f_sh[', data$stand_idxs[t],',',x, ']'))
  util$plot_conn_pushforward_quantiles(samples, names, data$all_years,
                                       xlab="", ylab='Stand GP', 
                                       display_ylim=c(-8, 3), display_xlim = c(1980, 2010))
  abline(v = c(2002,1996), lty = 3, col="darkgrey")
}
``` 

The model seems to over-regularize the large negative shocks. And for the first tree, it compensates for the stand-level GP shock in 2022 with a positive shock. We obtain a $\tau_\text{outer}^\text{shock}$ rather small compared to what our broad prior would suggest:

```{r}
#| fig-width: 7
#| fig-height: 3
#| echo: false

par(mfcol = c(1,2), cex.lab = 0.9)
util$plot_expectand_pushforward(samples[['outer_tau_sck']], 50,
                                flim = c(0,5), 
                                display_name=expression(tau[shock]^outer))
plot(1, type = "n", xlab = "", ylab = "", xlim = c(0, 10), ylim = c(0, 0.12),
     xaxs = "i", yaxs = "i", bty = "n", yaxt = "n")  
xs <- seq(0, 10, 0.01)
ys <- dnorm(xs, 0, 10 / 2.57)
lines(xs, ys, lwd=2.7, col='white')
lines(xs, ys, lwd=1.7, col=util$c_mid_teal)
``` 

# Considering a heterogeneous [$\omega_\text{shock}^\text{concordant}$]{style="text-transform:none;"}

We add some flexibility to our model! Now, we consider that:

- $\phi_\text{shock}$ and $\omega_\text{shock}^\text{concordant}$ varies across stands (heterogeneous)   

```{r}
if(FALSE){
  fit2 <- stan(file=file.path(wd, 'model', 'stan', 'model9_marginalshocks_zerotauinner_censoring_onlyconcordant_heterogeneous.stan'),
              data=data, seed=5838293, chains = 4,
              warmup=1000, iter=2024, refresh=10)
  saveRDS(fit2, file.path(wd, 'model/shocks/output/marginal', 'fit_normal_censoring_heterogeneous.rds'))
}else{
  fit2 <- readRDS(file.path(wd, 'model/shocks/output/marginal', 'fit_normal_censoring_heterogeneous.rds'))
}
``` 

Still a good amount of warnings! But only for one chain now.

```{r}
diagnostics <- util$extract_hmc_diagnostics(fit2)
util$check_all_hmc_diagnostics(diagnostics)
```

However the parameter-level warnings are more concerning.
```{r}
samples2 <- util$extract_expectand_vals(fit2)
base_samples <- util$filter_expectands(samples2,
                                       c('alpha', 
                                         'beta_gdd', 'beta_sm', 'beta_vpd',
                                         'rho_sh', 'gamma_sh',
                                         'rho_sp', 'gamma_sp',
                                         'sigma', 
                                         'phi_sck', 'omega_conc_sck'),
                                       check_arrays = TRUE)
util$check_all_expectand_diagnostics(base_samples)
``` 

We can look at the pair plot between $\sigma$ and $\gamma_\text{short}$:
```{r}
#| fig-width: 5.5
#| fig-height: 5.5
#| echo: false

util$plot_pairs_by_chain(samples2[['sigma']], expression(sigma),
                         samples2[['gamma_sh']], expression(gamma[short]))
``` 

Looking at the two stand GPs reveals the two modes for $\gamma_\text{short}$... at least on stand 1.
```{r}
#| fig-width: 8
#| fig-height: 3.5
#| echo: false

par(mfrow= c(1,2))
s <- 1
names <- sapply(1:data$N_all_years,
                  function(x) paste0('f_sh[', s,',',x, ']'))
util$plot_realizations(samples2, names, data$all_years,
                                       xlab="", ylab='Stand GP', 
                                       display_ylim=c(-8, 3), display_xlim = c(1980, 2010))

s <- 2
names <- sapply(1:data$N_all_years,
                  function(x) paste0('f_sh[', s,',',x, ']'))
util$plot_realizations(samples2, names, data$all_years,
                                       xlab="", ylab='Stand GP', 
                                       display_ylim=c(-8, 3), display_xlim = c(1980, 2010))
``` 

If we look at the pair plot between $\phi_\text{shock}$ and $\omega_\text{shock}^\text{concordant}$, for stand 1:
```{r}
#| fig-width: 5.5
#| fig-height: 5.5
#| echo: false

util$plot_pairs_by_chain(samples2[['phi_sck[1]']], expression(sigma),
                         samples2[['omega_conc_sck[1]']], expression(gamma[short]))

``` 

And stand 2:
```{r}
#| fig-width: 5.5
#| fig-height: 5.5
#| echo: false

util$plot_pairs_by_chain(samples2[['phi_sck[2]']], expression(sigma),
                         samples2[['omega_conc_sck[2]']], expression(gamma[short]))
``` 

However, the retrodictive check on the trees of stand 1 seems to indicate that we are on a good path!

```{r}
#| fig-width: 8
#| fig-height: 9.4
#| echo: false

par(mfcol = c(4,2), cex.lab = 1.5)
for(t in 1:4){
  idxs_t <- data$tree_start_idxs[t]:data$tree_end_idxs[t]
  names <- sapply(idxs_t,
                  function(x) paste0('log_rw_pred[', x, ']'))
  util$plot_conn_pushforward_quantiles(samples2, names, data$years[idxs_t],
                                       xlab="", ylab='Predictions (log)', main = paste0('Tree ', data$uniq_tree_ids[t]),
                                       display_ylim=c(-8, 2), display_xlim = c(1980, 2010))
  points(data$years[idxs_t], ifelse(data$log_rw_obs[idxs_t] == log(1e-4), NA, data$log_rw_obs[idxs_t]), pch=16, cex=1, col="white")
  points(data$years[idxs_t], ifelse(data$log_rw_obs[idxs_t] == log(1e-4), NA, data$log_rw_obs[idxs_t]), pch=16, cex=0.6, col="black")
  abline(v =  c(2002,1996), lty = 3, col="darkgrey")
  
  names <- sapply(idxs_t,
                  function(x) paste0('sck_state[', x, ']'))
  ticks <- c(unlist(lapply(seq(1980, 2000,10), function(x) c(x, rep(NA, 9)))), '2010')
  util$plot_disc_pushforward_quantiles(samples2, names, data$years[idxs_t],
                                       xlab="", ylab='Shock state', 
                                       display_ylim=c(-0.1, 1.1), 
                                       xticklabs =  ticks)
  abline(v =  c(2002,1996), lty = 3, col="darkgrey")
  
  names <- sapply(idxs_t,
                  function(x) paste0('delta_sck[', x, ']'))
  util$plot_conn_pushforward_quantiles(samples2, names, data$years[idxs_t],
                                       xlab="", ylab=expression(delta[shock]), 
                                       display_ylim=c(-8, 3), display_xlim = c(1980, 2010))
  abline(v =  c(2002,1996), lty = 3, col="darkgrey")
  
  names <- sapply(1:data$N_all_years,
                  function(x) paste0('f_sh[', data$stand_idxs[t],',',x, ']'))
  util$plot_conn_pushforward_quantiles(samples2, names, data$all_years,
                                       xlab="", ylab='Stand GP', 
                                       display_ylim=c(-8, 3), display_xlim = c(1980, 2010))
  abline(v = c(2002,1996), lty = 3, col="darkgrey")
}
``` 

```{r}
#| fig-width: 8
#| fig-height: 9.4

par(mfcol = c(4,2))
for(t in 41:44){
  idxs_t <- data$tree_start_idxs[t]:data$tree_end_idxs[t]
  names <- sapply(idxs_t,
                  function(x) paste0('log_rw_pred[', x, ']'))
  util$plot_conn_pushforward_quantiles(samples2, names, data$years[idxs_t],
                                       xlab="", ylab='log_rw_pred', main = paste0('Tree ', data$uniq_tree_ids[t]),
                                       display_ylim=c(-10, 2), display_xlim = c(1980, 2022))
  points(data$years[idxs_t], data$log_rw_obs[idxs_t], pch=16, cex=1, col="white")
  points(data$years[idxs_t], data$log_rw_obs[idxs_t], pch=16, cex=0.6, col="black")
  abline(v =  c(2002,2006), lty = 3, col="darkgrey")
  
  names <- sapply(idxs_t,
                  function(x) paste0('sck_state[', x, ']'))
  ticks <- c(unlist(lapply(seq(1980, 2010,10), function(x) c(x, rep(NA, 9)))), '2020',NA,NA)
  util$plot_disc_pushforward_quantiles(samples2, names, data$years[idxs_t],
                                       xlab="", ylab='Shock state', 
                                       display_ylim=c(-0.1, 1.1), 
                                       xticklabs =  ticks)
  abline(v =  c(2002,2006), lty = 3, col="darkgrey")
  
  names <- sapply(idxs_t,
                  function(x) paste0('delta_sck[', x, ']'))
  util$plot_conn_pushforward_quantiles(samples2, names, data$years[idxs_t],
                                       xlab="", ylab=expression(delta[shock]), 
                                       display_ylim=c(-10, 2), display_xlim = c(1980, 2022))
  abline(v =  c(2002,2006), lty = 3, col="darkgrey")
  
  names <- sapply(1:data$N_all_years,
                  function(x) paste0('f_sh[', data$stand_idxs[t],',',x, ']'))
  util$plot_conn_pushforward_quantiles(samples2, names, data$all_years,
                                       xlab="", ylab='Stand GP', 
                                       display_ylim=c(-2, 2), display_xlim = c(1980, 2022))
  abline(v = c(2002,2006), lty = 3, col="darkgrey")
}
```