---
title: "Marginalizing shocks"
toc: true
number-sections: true
highlight: pygments
pdf-engine: lualatex
format:
  html:
    html-math-method: katex
    theme:
      - lux
      - custom.scss
    standalone: true
    embed-resources: true
    code-overflow: wrap
    linkcolor: "#B97C7C"
    fontsize: '11pt'
    code-block-border-left: "#8F2727"
    code-block-bg: "#f8f8f8"
    highlight-style: github
  pdf:
    keep-tex: true
    fig-width: 4
    fig-height: 4
    code-overflow: wrap
    monofont: 'Latin Modern Mono'
    monofontoptions:
      - Scale=0.5
format-links: false
knitr:
  opts_chunk:
    comment: ''
    cache.lazy: FALSE
  opts_knit:
    global.par: TRUE
geometry:
  - top=20mm
  - left=20mm
  - bottom=20mm
  - right=20mm
---

```{css, echo=FALSE}
.cell-output { border-left: #B97C7C 2px solid; border-radius: 6px;}
```

*Cover that shocks which I ought not to see.* 

```{r}
#| echo: false
#| warning: false
#| message: false
wd <- "~/projects/climategrowthshifts/analysis/pnwandmore"
setwd(file.path(wd, 'model/shocks'))
util <- new.env()
source('mcmc_analysis_tools_rstan.R', local=util)
source('mcmc_visualization_tools.R', local=util)
# source('functions/custom_functions.R', local=util)
setwd(wd)
library(rstan)
options(mc.cores = parallel::detectCores())

chunk_hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- chunk_hook(x, options)
  paste0("\\linespread{0.5}\n", x, "\n\n\\linespread{2}")
})

par(family="serif", las=1, bty="l",
    cex.axis=0.8, cex.lab=0.85, cex.main=1,
    xaxs="i", yaxs="i", mar = c(5, 5, 3, 1))

logit <- function(x) log(x/(1-x))
inv_logit <- function(x) return(exp(x) / (1 + exp(x)))
```

# Model reparameterization

We marginalize the shocks in the mixture model!
Going from:
```{stan output.var = 'foo'}
#| eval: false
# Non-shocky
target += normal_lpdf(y[n] | mu[n] + delta_sck[n], sigma)
target += normal_lpdf(delta_sck[n] | 0, inner_tau_sck)

# Shocky
target += normal_lpdf(y[n] | mu[n] + delta_sck[n], sigma)
target += normal_lpdf(delta_sck[n] | 0, outer_tau_sck)
```

to a log-likelihood where shocks are marginalized out:
```{stan output.var = 'foo'}
#| eval: false
# Non-shocky
target += normal_lpdf(y[n] | mu[n], eta)

# Shocky
target += normal_lpdf( y[n] | mu[n], eta * sqrt{1 + nu / eta^2})
```

where:
$$\eta = \sqrt{ \sigma^2 + {\tau_\text{inner}^\text{shock}}^2 }$$
$$\nu = {\tau_\text{outer}^\text{shock}}^2-{\tau_\text{inner}^\text{shock}}^2$$
In Stan, we now have in the model block:

```{r}
writeLines(
  readLines(file.path(wd, 'model', 'stan', 
                      'model9_marginalshocks.stan'))[230:315])
```

# But... my shocks...

Our large shocks are likely tied to interesting ecological stuff... But by marginalizing them, we don't have inferences on them anymore. \
It could be doable to "*to marginalize out $\delta_\text{shock}$ in the “small shock” branch of the mixture model but leave $\delta_\text{shock}$ explicit in the “large shock” branch of the mixture model*" (Mike). \
Since I'm not totally sure how to do that, I decided to sample the $\delta_\text{shock}$ afterwards, to recover their posterior distributions in the generated quantities block. This is an ongoing project, so for now I recover only the shock state ($z_t$).

```{r}
writeLines(
  readLines(file.path(wd, 'model', 'stan', 
                      'model9_marginalshocks.stan'))[351:405])
```

# Test with a small dataset

Let's run this new model on a very small dataset, 21 trees and 1 site. \
The good thing here is that I don't need to tune the (non-)centered parametrization anymore.

```{r}
data <- readRDS(file = file.path(wd, 'output', 'model', 'data_15nov2025_az592.rds'))
data$N_stand_trees <- array(data$N_stand_trees, dim = 1)
data$N_stand_years <- array(data$N_stand_years, dim = 1)
data$stand_start_years_idxs <- array(data$stand_start_years_idxs, dim = 1)
data$stand_trees_end_idxs <- array(data$stand_trees_end_idxs, dim = 1)
data$stand_trees_start_idxs <- array(data$stand_trees_start_idxs, dim = 1)

if(FALSE){
  fit <- stan(file=file.path(wd, 'model', 'stan', 'model9_marginalshocks.stan'),
            data=data, seed=5838293, chains = 4,
            warmup=1000, iter=2024, refresh=10)
  saveRDS(fit, file.path(wd, 'model/shocks/output/marginal', 'fit.rds'))
}else{
  fit <- readRDS(file.path(wd, 'model/shocks/output/marginal', 'fit.rds'))
}

```

Let's look at the diagnostics:

```{r}
diagnostics <- util$extract_hmc_diagnostics(fit)
util$check_all_hmc_diagnostics(diagnostics)

samples <- util$extract_expectand_vals(fit)
base_samples <- util$filter_expectands(samples,
                                       c('alpha', 
                                         'beta_gdd', 'beta_sm', 'beta_vpd',
                                         'rho_sh', 'gamma_sh',
                                         'rho_sp', 'gamma_sp',
                                         'eta', 'nu',
                                         'phi_sck', 'omega_conc_sck', 'omega_nonconc_sck'),
                                       check_arrays = TRUE)
util$check_all_expectand_diagnostics(base_samples)

```

Clean! Let's continue. \
These are the latent shock states:

```{r}
#| fig-width: 7
#| fig-height: 4
set.seed(123456)
random_trees <- sample(1:data$N_trees, 4)

par(mfrow = c(2,2), mar = c(2, 5, 3, 1))
for(t in random_trees){
  idxs_t <- data$tree_start_idxs[t]:data$tree_end_idxs[t]
  names <- sapply(idxs_t,
                  function(x) paste0('sck_state[', x, ']'))
  util$plot_conn_pushforward_quantiles(samples, names, data$years[idxs_t],
                                       xlab="", ylab=paste0("Shock state - tree ",t), 
                                       display_ylim=c(-0.1, 1.1), display_xlim = c(1980, 2010))
}

```

# Next step

Is it possible to recover the posterior of $\delta_\text{shock}(k) ~|~ y(k)$, without having ${\tau_\text{inner}^\text{shock}}$ or ${\tau_\text{outer}^\text{shock}}$? \
Maybe a first step could be to fix: ${\tau_\text{inner}^\text{shock}} = 0$?

Because for now my predictions look like:

```{r}
#| fig-width: 7
#| fig-height: 4


par(mfrow = c(2,2), mar = c(2, 5, 3, 1))
for(t in random_trees){
  idxs_t <- data$tree_start_idxs[t]:data$tree_end_idxs[t]
  names <- sapply(idxs_t,
                  function(x) paste0('log_rw_pred[', x, ']'))
  util$plot_conn_pushforward_quantiles(samples, names, data$years[idxs_t],
                                       xlab="", ylab=paste0("Shock state - tree ",t), 
                                       display_ylim=c(-10, 5), display_xlim = c(1980, 2010))
  points(data$years[idxs_t], data$log_rw_obs[idxs_t], pch=16, cex=0.7, col="white")
  points(data$years[idxs_t], data$log_rw_obs[idxs_t], pch=16, cex=0.4, col="black")
}

```

# Assuming that [${\tau_\text{inner}^\text{shock}} = 0$]{style="text-transform:none;"}

::: {.callout-note}
Update Nov. 26 10pm: I tried another method in the section 8 to marginalize only the non-shocks, and reconstruct _a posteriori_ $\delta_\text{shock}$ that are $\neq 0$ only for large shocks. This does not use the normal-normal conjugancy anymore.
:::

If I assume that ${\tau_\text{inner}^\text{shock}} = 0$, I can reconstruct the large shocks using the normal-normal conjugancy (*merci* Google):
$$\delta_\text{shock}(k) ~|~ y_k \sim \text{normal} \left( ({\tau_\text{outer}^\text{shock}}^2 / ({\tau_\text{outer}^\text{shock}}^2 + \sigma^2)) \times y_k ~,~ \sqrt{({\tau_\text{outer}^\text{shock}}^2 \times \sigma^2) / ({\tau_\text{outer}^\text{shock}}^2 + \sigma^2)}\right)$$

In the generated quantities block, we can then have:

```{r}
writeLines(
  readLines(file.path(wd, 'model', 'stan', 
                      'model9_marginalshocks_zerotauinner.stan'))[398:405])
```


```{r}

if(FALSE){
  fit2 <- stan(file=file.path(wd, 'model', 'stan', 'model9_marginalshocks_zerotauinner.stan'),
            data=data, seed=5838293, chains = 4,
            warmup=1000, iter=2024, refresh=10)
  saveRDS(fit2, file.path(wd, 'model/shocks/output/marginal', 'fit2.rds'))
}else{
  fit2 <- readRDS(file.path(wd, 'model/shocks/output/marginal', 'fit2.rds'))
}

```

Diagnostics are cleaned:
```{r}
diagnostics <- util$extract_hmc_diagnostics(fit2)
util$check_all_hmc_diagnostics(diagnostics)

samples <- util$extract_expectand_vals(fit2)
base_samples <- util$filter_expectands(samples,
                                       c('alpha', 
                                         'beta_gdd', 'beta_sm', 'beta_vpd',
                                         'rho_sh', 'gamma_sh',
                                         'rho_sp', 'gamma_sp',
                                         'sigma', 'outer_tau_sck',
                                         'phi_sck', 'omega_conc_sck', 'omega_nonconc_sck'),
                                       check_arrays = TRUE)
util$check_all_expectand_diagnostics(base_samples)

```

We can look at the latent shocks:
```{r}
#| fig-width: 7
#| fig-height: 4


par(mfrow = c(2,2), mar = c(2, 5, 3, 1))
for(t in random_trees){
  idxs_t <- data$tree_start_idxs[t]:data$tree_end_idxs[t]
  names <- sapply(idxs_t,
                  function(x) paste0('delta_sck[', x, ']'))
  util$plot_conn_pushforward_quantiles(samples, names, data$years[idxs_t],
                                       xlab="", ylab=paste0("Delta shock - tree ",t), 
                                       display_ylim=c(-10, 0.1), display_xlim = c(1980, 2010))
}

```

And at the predictions:
```{r}
#| fig-width: 7
#| fig-height: 4


par(mfrow = c(2,2), mar = c(2, 5, 3, 1))
for(t in random_trees){
  idxs_t <- data$tree_start_idxs[t]:data$tree_end_idxs[t]
  names <- sapply(idxs_t,
                  function(x) paste0('log_rw_pred[', x, ']'))
  util$plot_conn_pushforward_quantiles(samples, names, data$years[idxs_t],
                                       xlab="", ylab=paste0("log(rw) - tree ",t), 
                                       display_ylim=c(-12, 1), display_xlim = c(1980, 2010))
   points(data$years[idxs_t], data$log_rw_obs[idxs_t], pch=16, cex=0.7, col="white")
  points(data$years[idxs_t], data$log_rw_obs[idxs_t], pch=16, cex=0.4, col="black")
}

```
\newpage
However, it seems that the shocks are bit too large...

```{r}
#| fig-width: 7
#| fig-height: 4


par(mfrow = c(2,2), mar = c(2, 5, 3, 1))
for(t in random_trees){
  idxs_t <- data$tree_start_idxs[t]:data$tree_end_idxs[t]
  names <- sapply(idxs_t,
                  function(x) paste0('log_rw_pred[', x, ']'))
  util$plot_conn_pushforward_quantiles(samples, names, data$years[idxs_t], 
                                       baseline_values = data$log_rw_obs[idxs_t],
                                       xlab="", ylab=paste0("Residuals - tree ",t), 
                                       display_ylim=c(-3, 1), display_xlim = c(1980, 2010),
                                       residual = TRUE)
}

```

And, let's not forget the stand-level GP, which is still wiggly:

```{r}
#| fig-width: 4
#| fig-height: 2.5

par(mfrow = c(1,1), mar = c(2, 5, 3, 1))
s <- 1
names <- sapply(1:data$N_all_years,
                  function(y) paste0('f_sh[', s, ',', y, ']'))
util$plot_realizations(samples, names, plot_xs = data$all_years[1:data$N_all_years], N_plots = 50,
                       main = paste0('Stand ', data$uniq_stand_ids[s]), ylab = 'f_sh',
                       display_xlim = data$all_years[c(1,data$N_all_years)], display_ylim = c(-2,2))
```

# More data!

Let's expand this stuff to moooore data! \
914 trees, this take more than 7 hours to run.

```{r}


data <- readRDS(file = file.path(wd, 'output/model', 'data_15nov2025_azPIPO.rds'))

if(FALSE){
  fit3 <- stan(file=file.path(wd, 'model', 'stan', 'model9_marginalshocks_zerotauinner.stan'),
            data=data, seed=5838293, chains = 4,
            warmup=1000, iter=2024, refresh=10)
  saveRDS(fit3, file.path(wd, 'model/shocks/output/marginal', 'fit3.rds'))
}else{
  fit3 <- readRDS(file.path(wd, 'model/shocks/output/marginal', 'fit3.rds'))
}

```

```{r}


diagnostics <- util$extract_hmc_diagnostics(fit3)
util$check_all_hmc_diagnostics(diagnostics)

samples <- util$extract_expectand_vals(fit3)
base_samples <- util$filter_expectands(samples,
                                       c('alpha', 
                                         'beta_gdd', 'beta_sm', 'beta_vpd',
                                         'rho_sh', 'gamma_sh',
                                         'rho_sp', 'gamma_sp',
                                         'sigma', 'outer_tau_sck',
                                         'phi_sck', 'omega_conc_sck', 'omega_nonconc_sck'),
                                       check_arrays = TRUE)
util$check_all_expectand_diagnostics(base_samples)

```

The model clearly wants to concentrate on lower values of $\omega_\text{shock}^\text{concordant}$. \

```{r}
#| fig-width: 7
#| fig-height: 7

util$plot_pairs_by_chain(samples[['omega_conc_sck']], 'omega_conc_sck',
                         samples[['omega_nonconc_sck']], 'omega_nonconc_sck')
```

If we look at all shock-related parameters, we can see that both modes for $\omega_\text{shock}^\text{concordant}$ are very far from our prior domain expertise.

```{r}
#| fig-width: 7
#| fig-height: 5

# Quickly look at posteriors
nf <- layout(
  matrix(c(1,1,1,2,2,2,
           3,3,4,4,5,5), 
         ncol=6, byrow=TRUE)
)
util$plot_expectand_pushforward(samples[['sigma']], 50,
                                flim = c(0,0.5), 
                                display_name=expression(tau[shock]^inner))
xs <- seq(0, 2, 0.01)
ys <- dnorm(xs, 0, log(1.1) / 2.57)
lines(xs, ys, lwd=2.7, col='white')
lines(xs, ys, lwd=1.7, col=util$c_mid_teal)

util$plot_expectand_pushforward(samples[['outer_tau_sck']], 50,
                                flim = c(0,10), 
                                display_name=expression(tau[shock]^outer))
xs <- seq(0, 10, 0.01)
ys <- dnorm(xs, 0, 10 / 2.57)
lines(xs, ys, lwd=2.7, col='white')
lines(xs, ys, lwd=1.7, col=util$c_mid_teal)

util$plot_expectand_pushforward(samples[['phi_sck[1]']], 50,
                                flim = c(0,0.5), 
                                display_name=expression(phi[schock]))
for(s in 2:data$N_stands){
  util$plot_expectand_pushforward(samples[[paste0('phi_sck[',s,']')]], 50,
                                  flim = c(0,0.5), 
                                  display_name=expression(phi[schock]), add = T)
}
xs <- seq(0, 0.5, 0.01)
ys <- dbeta(xs, 2, 20)
lines(xs, ys, lwd=2.7, col='white')
lines(xs, ys, lwd=1.7, col=util$c_mid_teal)

util$plot_expectand_pushforward(samples[['omega_conc_sck']], 50,
                                flim = c(0,1), 
                                display_name=expression(omega[shock]^concordant))
xs <- seq(0, 1, 0.01)
ys <- dbeta(xs, 230, 14)
lines(xs, ys, lwd=2.5, col='white')
lines(xs, ys, lwd=1.7, col=util$c_mid_teal)
util$plot_expectand_pushforward(samples[['omega_nonconc_sck']], 50,
                                flim = c(0,0.5), 
                                display_name=expression(omega[shock]^nonconcordant))
xs <- seq(0, 1, 0.01)
ys <- dbeta(xs, 1, 20)
lines(xs, ys, lwd=2.7, col='white')
lines(xs, ys, lwd=1.7, col=util$c_mid_teal)
```

# What are the concordant shocks captured by the model?

Let's dig into these tree-level shocks (*concordant*, i.e. when the stand is shocking) that are less frequent that we initially thought. \
We can first look at the stand we initially fitted alone (S14). Let's look at 10 random trees.

```{r}
#| fig-width: 7
#| fig-height: 12.5

trees <- which(grepl('az592' , data$uniq_tree_ids))
s1 <- unique(data$stand_idxs[trees])

random_trees <- sample(trees, 10)
par(mfrow = c(5,2), mar = c(2, 5, 3, 1))
for(t in random_trees){
  idxs_t <- data$tree_start_idxs[t]:data$tree_end_idxs[t]
  names <- sapply(idxs_t,
                  function(x) paste0('log_rw_pred[', x, ']'))
  util$plot_conn_pushforward_quantiles(samples, names, data$years[idxs_t],
                                       xlab="", ylab=paste0("log(rw) - tree ",t), 
                                       display_ylim=c(-12, 1), display_xlim = c(1980, 2010))
  points(data$years[idxs_t], data$log_rw_obs[idxs_t], pch=16, cex=0.7, col="white")
  points(data$years[idxs_t], data$log_rw_obs[idxs_t], pch=16, cex=0.4, col="black")
}
mtext(paste0("Stand ", data$uniq_stand_ids[s1]), side = 3, line = -2, outer = TRUE, adj = 0)


```

On this stand, it does seem that most trees are shocking. \
But on some other stand, the shock seem a bit less concordant across trees:

```{r}
#| fig-width: 7
#| fig-height: 12.5

s2 <- 6
trees <- which(data$stand_idxs == s2)
  
random_trees <- sample(trees, 10)
par(mfrow = c(5,2), mar = c(2, 5, 3, 1))
for(t in random_trees){
  idxs_t <- data$tree_start_idxs[t]:data$tree_end_idxs[t]
  names <- sapply(idxs_t,
                  function(x) paste0('log_rw_pred[', x, ']'))
  util$plot_conn_pushforward_quantiles(samples, names, data$years[idxs_t],
                                       xlab="", ylab=paste0("log(rw) - tree ",t), 
                                       display_ylim=c(-12, 1), display_xlim = c(1980, 2010))
   points(data$years[idxs_t], data$log_rw_obs[idxs_t], pch=16, cex=0.7, col="white")
  points(data$years[idxs_t], data$log_rw_obs[idxs_t], pch=16, cex=0.4, col="black")
}
mtext(paste0("Stand ", data$uniq_stand_ids[s2]), side = 3, line = -2, outer = TRUE, adj = 0)

```

The stand-level probability of shocks is not more important on the stand where the tree shocks are more concordant...

```{r}
#| fig-width: 4
#| fig-height: 2.5

par(mfrow = c(1,1))
util$plot_expectand_pushforward(samples[[paste0('phi_sck[',s1,']')]], 
                                50, flim = c(0,1), ylim = c(0,20),
                                display_name=expression(phi[shock]))
text(data$uniq_stand_ids[s1], x = 0.05, y = 16, col = util$c_dark)
util$plot_expectand_pushforward(samples[[paste0('phi_sck[',s2,']')]], 
                                50, flim = c(0,1), ylim = c(0,20),
                                display_name=expression(phi[shock]), add = TRUE, 
                                col = util$c_light_highlight)
text(data$uniq_stand_ids[s2], x = 0.17, y = 8, col = util$c_light_highlight)
```

Let's look at another stand!

```{r}
#| fig-width: 7
#| fig-height: 12.5

s3 <- 20
trees <- which(data$stand_idxs == s3)
  
random_trees <- sample(trees, 10)
par(mfrow = c(5,2), mar = c(2, 5, 3, 1))
for(t in random_trees){
  idxs_t <- data$tree_start_idxs[t]:data$tree_end_idxs[t]
  names <- sapply(idxs_t,
                  function(x) paste0('log_rw_pred[', x, ']'))
  util$plot_conn_pushforward_quantiles(samples, names, data$years[idxs_t],
                                       xlab="", ylab=paste0("log(rw) - tree ",t), 
                                       display_ylim=c(-12, 1), display_xlim = c(1980, 2010))
  abline(v = 2002, lty = 2, col = 'grey40')
  points(data$years[idxs_t], data$log_rw_obs[idxs_t], pch=16, cex=0.7, col="white")
  points(data$years[idxs_t], data$log_rw_obs[idxs_t], pch=16, cex=0.4, col="black")
}
mtext(paste0("Stand ", data$uniq_stand_ids[s3]), side = 3, line = -2, outer = TRUE, adj = 0)

```

On this stand, the trees seem to shock concordantly in 2002. But, is this low growth year considered as a shock by the model?

```{r}
#| fig-width: 7
#| fig-height: 12.5

par(mfrow = c(5,2), mar = c(2, 5, 3, 1))
for(t in random_trees){
  idxs_t <- data$tree_start_idxs[t]:data$tree_end_idxs[t]
  names <- sapply(idxs_t,
                  function(x) paste0('delta_sck[', x, ']'))
  util$plot_conn_pushforward_quantiles(samples, names, data$years[idxs_t],
                                       xlab="", ylab=paste0("log(rw) - tree ",t), 
                                       display_ylim=c(-12, 1), display_xlim = c(1980, 2010))
  abline(v = 2002, lty = 2, col = 'grey40')
  points(data$years[idxs_t], data$log_rw_obs[idxs_t], pch=16, cex=0.7, col="white")
  points(data$years[idxs_t], data$log_rw_obs[idxs_t], pch=16, cex=0.4, col="black")
}
mtext(paste0("Stand ", data$uniq_stand_ids[s3]), side = 3, line = -2, outer = TRUE, adj = 0)

```

Actually, no! The concordance is here captured by the stand-level GP.

```{r}
#| fig-width: 4.5
#| fig-height: 3

par(mfrow = c(1,1), mar = c(2, 5, 3, 1))
names <- sapply(1:31,
                  function(y) paste0('f_sh[', s3, ',', y, ']'))
util$plot_realizations(samples, names, plot_xs = data$all_years[1:31], N_plots = 50,
                       main = paste0('Stand ', data$uniq_stand_ids[s]), ylab = 'f_sh',
                       display_xlim = data$all_years[c(1,31)], display_ylim = c(-2,2))
abline(v = 2002, lty = 2, col = 'grey40')
```

I think that the shock model is trying only to capture the big shocks (the tail of distribution ?), and thus ignore the medium shocks. \
How to improve this?

# A new implementation to keep only the large shocks

::: {.callout-note}
Update Nov. 26 10pm: I try a new method to marginalize only the non-shocks, that does not use the normal-normal conjugancy anymore. This will be useful if we want to use another distribution with a heavier tail for the $\delta_\text{shock}$.
:::

What if I leave the $\delta_\text{shock}$ explicit in the large shock branch of the mixture model? \
I can set:
$$\delta_\text{large shock} \sim \text{normal}(0, {\tau_\text{outer}^\text{shock}})$$
So, for observations that end up in the non-shocky branch of the model, $\delta_\text{large shock}$ will be inform only by this prior... But then in the generated quantities block I can reconstruct $\delta_\text{shock}^\text{post}$, which is basically $=0$ in a non-shock year and $\delta_\text{large shock}$ in a shock year?

Note that in this case I also assume that $\tau_\text{inner}^\text{shock} = 0$.
So I have this in the model block:
```{r}
writeLines(
  readLines(file.path(wd, 'model', 'stan', 
                      'model9_marginalsmallshocks_explicitbigshocks.stan'))[295])

writeLines(
  readLines(file.path(wd, 'model', 'stan', 
                      'model9_marginalsmallshocks_explicitbigshocks.stan'))[308:318])
```
And in the generated quantities block:
```{r}
writeLines(
  readLines(file.path(wd, 'model', 'stan', 
                      'model9_marginalsmallshocks_explicitbigshocks.stan'))[374:379])
```

I run this model one the small dataset:

```{r}
data <- readRDS(file = file.path(wd, 'output', 'model', 'data_15nov2025_az592.rds'))
data$N_stand_trees <- array(data$N_stand_trees, dim = 1)
data$N_stand_years <- array(data$N_stand_years, dim = 1)
data$stand_start_years_idxs <- array(data$stand_start_years_idxs, dim = 1)
data$stand_trees_end_idxs <- array(data$stand_trees_end_idxs, dim = 1)
data$stand_trees_start_idxs <- array(data$stand_trees_start_idxs, dim = 1)

if(FALSE){
  fit2bis <- stan(file=file.path(wd, 'model', 'stan', 'model9_marginalsmallshocks_explicitbigshocks.stan'),
            data=data, seed=5838293, chains = 4,
            warmup=1000, iter=2024, refresh=10)
  saveRDS(fit2bis, file.path(wd, 'model/shocks/output/marginal', 'fit2bis.rds'))
}else{
  fit2bis <- readRDS(file.path(wd, 'model/shocks/output/marginal', 'fit2bis.rds'))
}

```

Diagnostics are not tooootally clean...

```{r}

diagnostics <- util$extract_hmc_diagnostics(fit2bis)
util$check_all_hmc_diagnostics(diagnostics)

samples <- util$extract_expectand_vals(fit2bis)
base_samples <- util$filter_expectands(samples,
                                       c('alpha', 
                                         'beta_gdd', 'beta_sm', 'beta_vpd',
                                         'rho_sh', 'gamma_sh',
                                         'rho_sp', 'gamma_sp',
                                         'sigma', 'outer_tau_sck',
                                         'phi_sck', 'omega_conc_sck', 'omega_nonconc_sck'),
                                       check_arrays = TRUE)
util$check_all_expectand_diagnostics(base_samples)

```

But let's still give a look at the latent $\delta_\text{large shock}$:
```{r}
#| fig-width: 7
#| fig-height: 4

set.seed(123456)
random_trees <- sample(1:data$N_trees, 4)

par(mfrow = c(2,2), mar = c(2, 5, 3, 1))
for(t in random_trees){
  idxs_t <- data$tree_start_idxs[t]:data$tree_end_idxs[t]
  names <- sapply(idxs_t,
                  function(x) paste0('delta_large_sck[', x, ']'))
  util$plot_conn_pushforward_quantiles(samples, names, data$years[idxs_t],
                                       xlab="", ylab=expression(delta[largeshock]), 
                                       display_ylim=c(-10, 10), display_xlim = c(1980, 2010))
}

```

As expected, their posteriors are informed by the data mostly in 2002 (the main shocky year).\
But if we look at the reconstructed $\delta_\text{shock}^\text{post}$:
```{r}
#| fig-width: 7
#| fig-height: 4


par(mfrow = c(2,2), mar = c(2, 5, 3, 1))
for(t in random_trees){
  idxs_t <- data$tree_start_idxs[t]:data$tree_end_idxs[t]
  names <- sapply(idxs_t,
                  function(x) paste0('delta_sck_post[', x, ']'))
  util$plot_conn_pushforward_quantiles(samples, names, data$years[idxs_t],
                                       xlab="", ylab=expression(delta[shock]^post), 
                                       display_ylim=c(-12, 1), display_xlim = c(1980, 2010))
}

```

And the corresponding predictions:
```{r}
#| fig-width: 7
#| fig-height: 4


par(mfrow = c(2,2), mar = c(2, 5, 3, 1))
for(t in random_trees){
  idxs_t <- data$tree_start_idxs[t]:data$tree_end_idxs[t]
  names <- sapply(idxs_t,
                  function(x) paste0('log_rw_pred[', x, ']'))
  util$plot_conn_pushforward_quantiles(samples, names, data$years[idxs_t],
                                       xlab="", ylab='log(rw)', 
                                       display_ylim=c(-12, 1), display_xlim = c(1980, 2010))
  points(data$years[idxs_t], data$log_rw_obs[idxs_t], pch=16, cex=0.7, col="white")
  points(data$years[idxs_t], data$log_rw_obs[idxs_t], pch=16, cex=0.4, col="black")
}

```

And the residuals:
```{r}
#| fig-width: 7
#| fig-height: 4


par(mfrow = c(2,2), mar = c(2, 5, 3, 1))
for(t in random_trees){
  idxs_t <- data$tree_start_idxs[t]:data$tree_end_idxs[t]
  names <- sapply(idxs_t,
                  function(x) paste0('log_rw_pred[', x, ']'))
  util$plot_conn_pushforward_quantiles(samples, names, data$years[idxs_t], 
                                       baseline_values = data$log_rw_obs[idxs_t],
                                       xlab="", ylab=paste0("Residuals - tree ",t), 
                                       display_ylim=c(-3, 1), display_xlim = c(1980, 2010),
                                       residual = TRUE)
}

```

We don't have the same issue that I had before with the shocks being a bit too large.
I think I was doing something wrong with the normal-normal conjugancy?
Likely because $\sigma$ is not the variability added to $\delta_\text{shock}$, but to $\mu + f + \delta_\text{shock}$.\
So when I was doing this:
```{r}
writeLines(
  readLines(file.path(wd, 'model', 'stan', 
                      'model9_marginalshocks_zerotauinner.stan'))[398:406])
```

Maybe the conjugate mean was wrong? I was conditioning on the observation `log_rw_obs[tree_idxs[y]]`, but maybe I should have conditioned on `(log_rw_obs[tree_idxs[y]] - mu[y] - f[tree_idxs[y]])`, i.e. rather do this:
```{r}
writeLines(
  readLines(file.path(wd, 'model', 'stan', 
                      'model9_marginalshocks_zerotauinner_updated.stan'))[398:407])
```
I did not try, though.

# Exploring the shocks that are not captured

We've seen that the shock model is able to capture _only_ the large shocks (missing rings), and let the stand-level GP contort to get the medium ones. \
Trying to diagnostic a model that takes hours to run is a pain, so I subset the data to two of the sites I mentioned before: one with large shocks (S14), and one with medium ones (S15).

```{r}
data <- readRDS(file = file.path(wd, 'output', 'model', 'data_15nov2025_az592_az628.rds'))

if(FALSE){
  fit5 <- stan(file=file.path(wd, 'model', 'stan', 'model9_marginalsmallshocks_explicitbigshocks_lessconcordant.stan'),
            data=data, seed=5838293, chains = 4,
            warmup=1000, iter=2024, refresh=10)
  saveRDS(fit5, file.path(wd, 'model/shocks/output/marginal', 'fit5.rds'))
}else{
  fit5 <- readRDS(file.path(wd, 'model/shocks/output/marginal', 'fit5.rds'))
}
```

As expected, we also get funnel-related warnings!

```{r}
diagnostics <- util$extract_hmc_diagnostics(fit5)
util$check_all_hmc_diagnostics(diagnostics)

samples <- util$extract_expectand_vals(fit5)
base_samples <- util$filter_expectands(samples,
                                       c('alpha', 
                                         'beta_gdd', 'beta_sm', 'beta_vpd',
                                         'rho_sh', 'gamma_sh',
                                         'rho_sp', 'gamma_sp',
                                         'sigma', 'outer_tau_sck',
                                         'phi_sck', 'omega_conc_sck', 'omega_nonconc_sck'),
                                       check_arrays = TRUE)
util$check_all_expectand_diagnostics(base_samples)
```

Notably, we get $\hat{R}$ warnings for $\tau_\text{inner}^\text{shock}$ and $\sigma$:
```{r}
#| fig-width: 6
#| fig-height: 6
util$plot_pairs_by_chain(samples[['sigma']], expression(sigma), samples[['outer_tau_sck']], expression(tau[shock]^outer))
```
And quite a lot of issues with our $\delta_\text{large shock}$:

```{r}
base_samples <- util$filter_expectands(samples,
                                       c('delta_large_sck'),
                                       check_arrays = TRUE)
util$check_all_expectand_diagnostics(base_samples)
```

Let's look at one of the shock that doesn't have a nice geometry:

```{r}
#| fig-width: 4
#| fig-height: 4
util$plot_div_pairs(paste0('delta_large_sck[', 1712, ']'), 'sigma', samples, diagnostics, transforms = list('sigma' = 1))
```

This shock corresponds to the dashed line below:

```{r}
#| fig-width: 7
#| fig-height: 7

par(mfrow = c(3,1))
t <- 46
idxs_t <- data$tree_start_idxs[t]:data$tree_end_idxs[t]
names <- sapply(idxs_t,
                function(x) paste0('log_rw_pred[', x, ']'))
util$plot_conn_pushforward_quantiles(samples, names, data$years[idxs_t],
                                     xlab="", ylab='log_rw_pred', 
                                     display_ylim=c(-2, 2), display_xlim = c(1980, 2021))
points(data$years[idxs_t], data$log_rw_obs[idxs_t], pch=16, cex=1, col="white")
points(data$years[idxs_t], data$log_rw_obs[idxs_t], pch=16, cex=0.6, col="black")
abline(v = data$years[1712], lty = 2)


names <- sapply(idxs_t,
                function(x) paste0('delta_sck_post[', x, ']'))
util$plot_conn_pushforward_quantiles(samples, names, data$years[idxs_t],
                                     xlab="", ylab='delta_sck_post', 
                                     display_ylim=c(-2, 2), display_xlim = c(1980, 2021))
abline(v = data$years[1712], lty = 2)

names <- sapply(1:43,
                function(x) paste0('f_sh[', data$stand_idxs[t],',',x, ']'))
util$plot_conn_pushforward_quantiles(samples, names, data$years[idxs_t],
                                     xlab="", ylab='Stand GP', 
                                     display_ylim=c(-2, 2), display_xlim = c(1980, 2021))
abline(v = data$years[1712], lty = 2)
```

As a side note, the diagnostics are cleaner when we keep everything marginalized and we reconstruct the $\delta_\text{shock}$ with the normal-normal conjugancy (corrected):

```{r}
data <- readRDS(file = file.path(wd, 'output', 'model', 'data_15nov2025_az592_az628.rds'))

if(FALSE){
  fit6 <- stan(file=file.path(wd, 'model', 'stan', 'model9_marginalshocks_zerotauinner_updated.stan'),
            data=data, seed=5838293, chains = 4,
            warmup=1000, iter=2024, refresh=10)
  saveRDS(fit6, file.path(wd, 'model/shocks/output/marginal', 'fit6.rds'))
}else{
  fit6 <- readRDS(file.path(wd, 'model/shocks/output/marginal', 'fit6.rds'))
}
```

```{r}
diagnostics <- util$extract_hmc_diagnostics(fit6)
util$check_all_hmc_diagnostics(diagnostics)

samples <- util$extract_expectand_vals(fit6)
base_samples <- util$filter_expectands(samples,
                                       c('alpha', 
                                         'beta_gdd', 'beta_sm', 'beta_vpd',
                                         'rho_sh', 'gamma_sh',
                                         'rho_sp', 'gamma_sp',
                                         'sigma', 'outer_tau_sck',
                                         'phi_sck', 'omega_conc_sck', 'omega_nonconc_sck'),
                                       check_arrays = TRUE)
util$check_all_expectand_diagnostics(base_samples)
```

And for the tree mentioned above we obtain:

```{r}
#| fig-width: 7
#| fig-height: 7

par(mfrow = c(3,1))
t <- 46
idxs_t <- data$tree_start_idxs[t]:data$tree_end_idxs[t]
names <- sapply(idxs_t,
                function(x) paste0('log_rw_pred[', x, ']'))
util$plot_conn_pushforward_quantiles(samples, names, data$years[idxs_t],
                                     xlab="", ylab='log_rw_pred', 
                                     display_ylim=c(-2, 2), display_xlim = c(1980, 2021))
points(data$years[idxs_t], data$log_rw_obs[idxs_t], pch=16, cex=1, col="white")
points(data$years[idxs_t], data$log_rw_obs[idxs_t], pch=16, cex=0.6, col="black")
abline(v = data$years[1712], lty = 2)


names <- sapply(idxs_t,
                function(x) paste0('delta_sck[', x, ']'))
util$plot_conn_pushforward_quantiles(samples, names, data$years[idxs_t],
                                     xlab="", ylab='delta_sck', 
                                     display_ylim=c(-2, 2), display_xlim = c(1980, 2021))
abline(v = data$years[1712], lty = 2)

names <- sapply(1:43,
                function(x) paste0('f_sh[', data$stand_idxs[t],',',x, ']'))
util$plot_conn_pushforward_quantiles(samples, names, data$years[idxs_t],
                                     xlab="", ylab='Stand GP', 
                                     display_ylim=c(-2, 2), display_xlim = c(1980, 2021))
abline(v = data$years[1712], lty = 2)
```

<!-- # Remove non-concordant shocks -->

<!-- There is a geometry issue that is likely due to the centered parametrization (I think?). \ -->
<!-- This is not an issue with large shocks, that are really informative. For the non-shocky years, it's only the prior that plays. \ -->
<!-- Is this more an issue with intermediate shocks, where the balance between prior and likelihood is tricky? Anyway, I want to step back and allow only for concordant shocks, i.e fix  $\omega_\text{shock}^\text{non-concordant} = 0$ as it will likely help me diagnose what is going on. -->

<!-- ```{r} -->
<!-- data <- readRDS(file = file.path(wd, 'output', 'model', 'data_15nov2025_az592_az628.rds')) -->

<!-- if(FALSE){ -->
<!--   fit6 <- stan(file=file.path(wd, 'model', 'stan', 'model9_marginalsmallshocks_explicitbigshocks_onlyconcordant.stan'), -->
<!--             data=data, seed=5838293, chains = 4, -->
<!--             warmup=1000, iter=2024, refresh=10) -->
<!--   saveRDS(fit6, file.path(wd, 'model/shocks/output/marginal', 'fit6.rds')) -->
<!-- }else{ -->
<!--   fit6 <- readRDS(file.path(wd, 'model/shocks/output/marginal', 'fit6.rds')) -->
<!-- } -->
<!-- ``` -->

<!-- ```{r} -->
<!-- diagnostics <- util$extract_hmc_diagnostics(fit6) -->
<!-- util$check_all_hmc_diagnostics(diagnostics) -->

<!-- samples <- util$extract_expectand_vals(fit6) -->
<!-- base_samples <- util$filter_expectands(samples, -->
<!--                                        c('alpha',  -->
<!--                                          'beta_gdd', 'beta_sm', 'beta_vpd', -->
<!--                                          'rho_sh', 'gamma_sh', -->
<!--                                          'rho_sp', 'gamma_sp', -->
<!--                                          'sigma', 'outer_tau_sck', -->
<!--                                          'phi_sck', 'omega_conc_sck'), -->
<!--                                        check_arrays = TRUE) -->
<!-- util$check_all_expectand_diagnostics(base_samples) -->
<!-- ``` -->

<!-- What's happening with the shock we were previously looking at: -->

<!-- ```{r} -->
<!-- #| fig-width: 4 -->
<!-- #| fig-height: 4 -->
<!-- util$plot_div_pairs(paste0('delta_large_sck[', 1712, ']'), 'outer_tau_sck', samples, diagnostics, transforms = list('outer_tau_sck' = 1)) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- #| fig-width: 7 -->
<!-- #| fig-height: 7 -->

<!-- par(mfrow = c(3,1)) -->
<!-- t <- 46 -->
<!-- idxs_t <- data$tree_start_idxs[t]:data$tree_end_idxs[t] -->
<!-- names <- sapply(idxs_t, -->
<!--                 function(x) paste0('log_rw_pred[', x, ']')) -->
<!-- util$plot_conn_pushforward_quantiles(samples, names, data$years[idxs_t], -->
<!--                                      xlab="", ylab='log_rw_pred',  -->
<!--                                      display_ylim=c(-2, 2), display_xlim = c(1980, 2021)) -->
<!-- points(data$years[idxs_t], data$log_rw_obs[idxs_t], pch=16, cex=1, col="white") -->
<!-- points(data$years[idxs_t], data$log_rw_obs[idxs_t], pch=16, cex=0.6, col="black") -->
<!-- abline(v = data$years[1712], lty = 2) -->


<!-- names <- sapply(idxs_t, -->
<!--                 function(x) paste0('delta_sck_post[', x, ']')) -->
<!-- util$plot_conn_pushforward_quantiles(samples, names, data$years[idxs_t], -->
<!--                                      xlab="", ylab='delta_sck_post',  -->
<!--                                      display_ylim=c(-2, 2), display_xlim = c(1980, 2021)) -->
<!-- abline(v = data$years[1712], lty = 2) -->

<!-- names <- sapply(1:43, -->
<!--                 function(x) paste0('f_sh[', data$stand_idxs[t],',',x, ']')) -->
<!-- util$plot_conn_pushforward_quantiles(samples, names, data$years[idxs_t], -->
<!--                                      xlab="", ylab='Stand GP',  -->
<!--                                      display_ylim=c(-2, 2), display_xlim = c(1980, 2021)) -->
<!-- abline(v = data$years[1712], lty = 2) -->
<!-- ``` -->


<!-- # What are concordant shocks? -->

<!-- Previous results clearly call for an update of our domain expertise. \ -->
<!-- What do we expect for  $\omega_\text{shock}^\text{concordant}$? Let's run the model with a less restricted prior. \ -->

<!-- ```{r} -->
<!-- #| fig-width: 4 -->
<!-- #| fig-height: 4 -->
<!-- #| echo: false -->

<!-- par(mfrow = c(1,1), mar = c(5, 5, 3, 1)) -->
<!-- plot(NULL, xlim=c(0,1), ylim=c(0,5), axes = FALSE, xlab = expression(omega[shock]^concordant), ylab ='') -->
<!-- axis(side = 1, at = seq(0,1,0.25)) -->
<!-- xs <- seq(0.75, 1, 0.01) -->
<!-- ys <- dbeta(xs, 230, 14) -->
<!-- lines(xs, ys, lwd=2.7, col='white') -->
<!-- lines(xs, ys, lwd=1.7, col=util$c_mid_teal, lty = 2) -->
<!-- text(x = 0.77, y = 4, label = "Old prior", col=util$c_mid_teal) -->
<!-- xs <- seq(0, 1, 0.01) -->
<!-- ys <- dbeta(xs, 5.8, 2.7) -->
<!-- lines(xs, ys, lwd=3.7, col='white') -->
<!-- lines(xs, ys, lwd=2.7, col=util$c_mid) -->
<!-- text(x = 0.45, y = 2, label = "New prior", col=util$c_mid) -->
<!-- ``` -->

<!-- I also want to implement a partial pooling on the stand-specific $\phi_\text{shock}$. I consider a parameterization in terms of the log-odds $\alpha_\text{shock} = \text{logit}(\phi_\text{shock})$, so we can then use a simple normal hierarchical prior: -->
<!-- $$ \alpha_\text{shock} \sim \text{normal}(\mu_{\alpha_\text{shock}}, \tau_{\alpha_\text{shock}})$$ -->
<!-- We thus have to define hyperpriors on $\mu_{\alpha_\text{shock}}$ and $\tau_{\alpha_\text{shock}}$ that are coherent with the prior we initally set: -->
<!-- $$\phi_\text{shock} \sim \text{beta}(2,20)$$ -->

<!-- ```{r} -->
<!-- #| fig-width: 8 -->
<!-- #| fig-height: 4 -->
<!-- par(mfrow = c(1,2)) -->
<!-- phi_sck <- rbeta(1e6, 2, 20) -->
<!-- hist(phi_sck, yaxt='n', ylab = '', col=util$c_mid, border='white', main = '', breaks = 30, xlim = c(0,1)) -->
<!-- hist(logit(phi_sck), yaxt='n', ylab = '', col=util$c_mid, border='white', main = '', breaks = 50, xlim = c(-7,2)) -->

<!-- ``` -->

<!-- The distribution of $\text{logit}(\phi_\text{shock})$ is skewed, but as a rough approximation we will choose a normal hyperprior: -->
<!-- $$\mu_{\alpha_\text{shock}} \sim \text{normal}(-2.5,0.8)$$ -->
<!-- ```{r} -->
<!-- #| fig-width: 8 -->
<!-- #| fig-height: 4 -->

<!-- mu_alpha_sck <- rnorm(1e6, mean = -2.5, sd = 0.8) -->
<!-- hist(mu_alpha_sck, yaxt='n', ylab = '', col=util$c_mid, border='white', main = '',  breaks = 50, xlim = c(-7,2)) -->
<!-- hist(inv_logit(mu_alpha_sck), yaxt='n', ylab = '', col=util$c_mid, border='white', main = '', breaks = 30, xlim = c(0,1)) -->
<!-- ``` -->

<!-- For $\tau_{\alpha_\text{shock}}$, I don't have a really good domain expertise, so I will set $\text{normal}(0,1)$. If $\mu_{\alpha_\text{shock}} = -2.5$, we would obtain: -->

<!-- ```{r} -->
<!-- #| fig-width: 8 -->
<!-- #| fig-height: 4 -->

<!-- mu_alpha_sck <- -2.5 -->
<!-- tau <- abs(rnorm(1e6, 0, 1)) -->
<!-- alpha_sck <- rnorm(1e6, mean = mu_alpha_sck, sd = tau) -->
<!-- hist(inv_logit(alpha_sck), yaxt='n', ylab = '', col=util$c_mid, border='white', main = '', breaks = 50, xlim = c(0,1)) -->
<!-- # abline(v = inv_logit(mu_alpha_sck), col = 'white', lwd = 2) -->
<!-- # abline(v = inv_logit(mu_alpha_sck), col = util$c_dark_teal, lwd = 1) -->

<!-- ``` -->

<!-- Let's run the model! -->

<!-- ```{r} -->
<!-- #| eval: false -->

<!-- data <- readRDS(file = file.path(wd, 'output/model', 'data_15nov2025_azPIPO.rds')) -->

<!-- if(FALSE){ -->
<!--   fit4 <- stan(file=file.path(wd, 'model', 'stan', 'model9_marginalshocks_zerotauinner_lessconcordant_ppphi.stan'), -->
<!--             data=data, seed=5838293, chains = 4, -->
<!--             warmup=1000, iter=2024, refresh=10) -->
<!--   saveRDS(fit4, file.path(wd, 'model/shocks/output/marginal', 'fit4.rds')) -->
<!-- }else{ -->
<!--   fit4 <- readRDS(file.path(wd, 'model/shocks/output/marginal', 'fit4.rds')) -->
<!-- } -->

<!-- ``` -->



