
---
title: "Modelling tree growth with Gaussian processes"
author:   
  - Victor
  - Lizzie
date: "May 2025"
date-format: "MMMM YYYY"
toc: true
number-sections: true
highlight: pygments
execute:
  enabled: true
format:
  html:
    html-math-method: katex
    theme:
      - lux
    embed-resources: true
    code-overflow: wrap
    linkcolor: "#B97C7C"
    page-layout: full
    fig-width: 8
    fig-height: 4
  pdf:
    keep-tex: true
    fig-width: 5.5
    fig-height: 7.5
    code-overflow: wrap
    monofontoptions:
      - Scale=0.5
knitr:
  opts_chunk:
    comment: ''
  opts_knit:
    global.par: TRUE
format-links: false
---

```{r}
#| echo: true

library(terra)
library(leaflet)
library(ggplot2)

kippenberger <- c("#8B174DFF", "#AE2565FF", "#C1447EFF", "#D06C9BFF", "#DA9FB8FF", 
                  "#ADBE7CFF", "#8BA749FF", "#6E8537FF", "#4F5F28FF", "#343D1FFF")

wd <- "~/projects/climategrowthshifts/analysis/mountrainier"

```


## Gaussian process model?

Following our one-week workshop in Banff, we came up with a new model of tree growth---with a big contribution of Mike Betancourt. 
The model is based on Gaussian processes.

### A quick overlook of Gaussian processes

From a simplified (and likely eroneous?) point of view, a (finite-$\mathrm{N}$-dimensional projection) of a Gaussian process $f_\mathrm{N}$ is defined as:
$$f_\mathrm{N} \sim \text{multinormal}(\mathrm{m}, \mathrm{K})$$
with a certain mean function $\mathrm{m}$ and a covariance matrix $\mathrm{K}$.  
In particular, the elements of $\mathrm{K}$ are defined by a covariance function.   
In our case (and in most Gaussian process applications), we use an *exponentiated quadratic* function:  
$$\forall i,j  \in [\![1,\mathrm{N}]\!] \hspace{1cm} \mathrm{k}_{i,j} = {\gamma^2} \exp\left(-\frac{1}{2}\left(\frac{\lvert x_j-x_i \rvert }{\rho}\right)^2\right)$$
This covariance function capture the "link" between observations along a continuous variable, such as time or space.   
The function is characterized by two parameters:  
- a marginal deviation $\gamma$, which controls the variability around the mean (the intensity?)  
- a length scale $\rho$, which controls how long/far the correlations persist  

```{r}
#| fig-height: 4
#| fig-width: 5
#| echo: false

exp_quad <- function(dx, gamma, rho){
  return(gamma^2 * exp(-1/2*(dx/rho)^2))
}

plotdf <- data.frame(dist = seq(0, 50, 0.1), cov = NA)
gamma <- 2
rho <- 15
plotdf$cov <- sapply(plotdf$dist, function(i) exp_quad(i, gamma = gamma, rho = rho))


plot.new()
plot.window(xlim = c(0,max(plotdf$dist)), ylim = c(0,gamma^2))
grid()
axis(1, las = 1, cex.axis = 0.8, tck=-0.02, at = c(0,rho, 50), labels=c("0", expression(rho), ""))
axis(2, las = 1, cex.axis = 0.8, tck=-0.02, at = c(0,gamma^2/exp(1/2),gamma^2), labels=c("0", expression(frac(gamma^2,exp(1/2))), expression(gamma^2)))
title(main = "Covariogram", line=1, cex.main = 1)
title(xlab = expression(paste(Delta,x)),  line=3, cex.main = 1)
lines(plotdf$dist, plotdf$cov, col="white", lwd=4)
lines(plotdf$dist, plotdf$cov, col="#AE2565FF", lwd=2)

```

### Application to tree growth

Now, let's say that we have ringwidth measurements $r_w$ for one tree, for each year $x_n$.
Using one Gaussian process, a simple model would look like this:

$$\forall x_n  \hspace{0.5cm} \log{r_w(x_n)} \sim \text{normal}(\mu + f(x_n), \sigma^2)$$
with:
$$\vec{f} \sim \text{multinormal}(\vec{0}, \mathrm{K})$$
The two equations are equivalent to:

$$\log\vec{{\text{rw}}} \sim \text{multinormal}(\vec{\mu}, \hspace{0.08cm} \Sigma + \mathrm{K})$$
where:

$$
    \vec{\mu} = \begin{bmatrix}
               \mu \\
               \mu \\
               \vdots \\
               \mu
    \end{bmatrix}
    \hspace{1cm}
	\Sigma = 
	\begin{bmatrix}
		\sigma^2  & 0 & \ldots & 0 \\
		0 & \ \sigma^2  & \ldots  & \vdots\\
		\vdots & \vdots & \ddots & \vdots \\
		0 & \ldots &  \ldots & \sigma^2 
	\end{bmatrix}
$$
\

In practice, we choose to model the $\log$ of ringwidth with two Gaussian processes:  

* one **long-term scale** Gaussian process $f_{\text{long}}$, to capture the long-term variation of tree growth, likely related to allometric constraints   
* one **short-term scale** Gaussian process $f_{\text{short,p}}$ per stand $p$, to capture some climatic variations over several years

The long-term scale GP is characterized by one pair of parameters $\{\gamma_{\text{sp}}, \rho_{\text{sp}}\}$ per species $sp$.   
The short-term scale GP is characterized by two parameters $\{\gamma_{\text{short}}, \rho_{\text{short}}\}$, common across all stands. The short-term scale does not depend on species, but its effect is scaled by a factor $\kappa_\text{sp}$ (see below).

\

The full model is, for $n$ observations for one tree $k$ of the species $sp$ on the stand $p$:

$$\log\vec{\text{rw}}_{sp,p,k} \sim \text{multinormal}(\vec{\mu_{sp,p}}, \hspace{0.08cm} \Sigma + \mathrm{K_{sp}})$$
with:
$$
\log\vec{\text{rw}}_{sp,k} = 
\log \circ \begin{bmatrix}
  \text{rw}_{sp,p,k}(t_1) \\
  \text{rw}_{sp,p,k}(t_2) \\
  \vdots \\
  \text{rw}_{sp,k}(t_n)
\end{bmatrix}
\hspace{1cm}
\log\vec{\mu}_{sp,p} = 
\log \circ \begin{bmatrix}
  \mu_{sp,p}(t_1) \\
  \mu_{sp,p}(t_2) \\
  \vdots \\
  \mu_{sp,p}(t_n)
\end{bmatrix}
$$

$$ \forall i  \in [\![1,\mathrm{n}]\!]  \hspace{0.5cm} \mu_{sp,p}(t_i)= \alpha
    + \beta_{\text{GSL}} * \text{GSL}(p,t_i)
    + \beta_{\text{GDD}}  * \text{GDD}(p,t_i)
    + \beta_{\text{SM}}  * \text{SM}(p,t_i)
    + \kappa_{\text{sp}}
    * f_{\text{short,p}}(t_i)$$

$$\vec{f}_{\text{short,p}} \sim \text{multinormal}(\vec{0}, \mathrm{K_\text{short}})$$
\

We look at three different climate predictors, at the stand scale:

* the growing season length $\text{GSL}(p,t_i)$, in days
* the growing degree days $\text{GDD}(p,t_i)$ accumulated during the growing season, in $\text{x}10$ $^{\circ}$C
* the mean soil moisture $\text{SM}(p,t_i)$ (weighted average across soil horizons) during the growing season, in $\%$

::: {.callout-note}
To discuss: do we really need the growing season length?
:::

## Modeling tree growth on Mt Rainier

### Study area



We have 16 different plots:

```{r}
#| out-height: 400px
#| out-width: 400px
#| echo: false

plots <- read.csv(file.path(wd, "data", "treerings_ailene", "tree_plot_climate_temp.csv"))
plots_vect <- vect(plots[,c("Longitude", "Latitude")], geom=c("Longitude", "Latitude"), crs = "EPSG:4326")

leaflet(plots) %>% addTiles() %>%
  addCircleMarkers(~Longitude, ~Latitude, label = ~Plot,
                   color = 'white', 
                   stroke = FALSE,
                   radius = 8, 
                   fillOpacity = 1) %>%
  addCircleMarkers(~Longitude, ~Latitude,  label = ~Plot,
                   color = 'black', 
                   stroke = FALSE,
                   radius = 4, 
                   fillOpacity = 1)


```

### Climate predictors

```{r}

raw_data <- read.csv(file.path(wd, 'input/treerings/mosttrees_allspecies_allplots_19802008.csv'))
clim_pred <- readRDS(file.path(wd, "data", "climate","processed_predictors", "wldas_climpredictors.rds"))

clim_pred$soilmoist_ings <- clim_pred$soilmoist_ings*100 # in percentage
clim_pred$gdd_ings <- clim_pred$gdd_ings/10 # in 10degC

```

#### Soil moisture

As said before, we first investigate the effect of 3 different predictors.  
All of them were extracted from WLDAS (from 1980).
We can look at the soil moisture:  

```{r}
#| echo: false

ggplot() +
  geom_line(data = clim_pred, aes(x = as.Date(paste(year, 01, 01, sep = "-")), y = soilmoist_ings, color = ID, group = ID), linewidth = 0.5) +
  scale_color_gradientn(colors = kippenberger) + 
  theme_bw() +
  scale_x_date(expand = c(0,0), date_labels = "%Y", breaks =  as.Date(paste(seq(1980,2020,5), 01, 01, sep = "-"))) +
  theme(legend.position = 'none',
        panel.grid = element_blank()) +
  labs(x = "Year", y = "Mean soil moisture during GS (%)")

```

We observe some big difference across stands:

```{r}
#| out-height: 400px
#| out-width: 700px
#| echo: false

climplotdf <- unique(clim_pred[, c('ID', 'plotname', 'alt', 'meansm', 'meangdd')])
climplotdf <- merge(climplotdf, plots[, c("Plot", "Longitude", "Latitude")], by.x = 'plotname', by.y = "Plot")

pal <- colorNumeric(
  palette = "Blues",
  domain = climplotdf$meansm*100)

leaflet(climplotdf) %>% addTiles() %>%
  addCircleMarkers(~Longitude, ~Latitude, label = ~plotname,
                   color = 'black', 
                   stroke = FALSE,
                   radius = 8, 
                   fillOpacity = 1) %>%
  addCircleMarkers(~Longitude, ~Latitude,  label = ~plotname,
                   color =  ~pal(meansm*100), 
                   stroke = FALSE,
                   radius = 5, 
                   fillOpacity = 1) %>%
  addLegend("bottomright", pal = pal, values = ~meansm*100,
    title = "Soil moist. (%)",
    opacity = 1
  )

```

\ 

If we look at the anomaly from the average stand soil moisture, we observe some similar patterns:

```{r}
#| echo: false

ggplot() +
  geom_line(data = clim_pred, aes(x = as.Date(paste(year, 01, 01, sep = "-")), y = soilmoist_ings-meansm*100, color = ID, group = ID), linewidth = 0.5) +
  scale_color_gradientn(colors = kippenberger) +
  theme_bw() +
  scale_x_date(expand = c(0,0), date_labels = "%Y", breaks =  as.Date(paste(seq(1980,2020,5), 01, 01, sep = "-"))) +
  theme(legend.position = 'none',
        panel.grid = element_blank()) +
  labs(x = "Year", y = "Mean soil moisture during GS (anomaly, %)")

```

#### Growing degree days

We can do the same for GDD:

```{r}
#| echo: false

ggplot() +
  geom_line(data = clim_pred, aes(x = as.Date(paste(year, 01, 01, sep = "-")), y = gdd_ings, color = ID, group = ID), linewidth = 0.5) +
  scale_color_gradientn(colors = kippenberger) + 
  theme_bw() +
  scale_x_date(expand = c(0,0), date_labels = "%Y", breaks =  as.Date(paste(seq(1980,2020,5), 01, 01, sep = "-"))) +
  theme(legend.position = 'none',
        panel.grid = element_blank()) +
  labs(x = "Year", y = "GDD accumulated in GS (x10 degC)")

```

```{r}
#| out-height: 400px
#| out-width: 700px
#| echo: false

pal <- colorNumeric(
  palette = "Reds",
  domain = climplotdf$meangdd/10)

leaflet(climplotdf) %>% addTiles() %>%
  addCircleMarkers(~Longitude, ~Latitude, label = ~plotname,
                   color = 'black', 
                   stroke = FALSE,
                   radius = 8, 
                   fillOpacity = 1) %>%
  addCircleMarkers(~Longitude, ~Latitude,  label = ~plotname,
                   color =  ~pal(meangdd/10), 
                   stroke = FALSE,
                   radius = 5, 
                   fillOpacity = 1) %>%
  addLegend("bottomright", pal = pal, values = ~meangdd/10,
    title = "GDD (x10 degC)",
    opacity = 1
  )

```

If we look at the anomaly from the average stand GDD, we also observe similar patterns:

```{r}
#| echo: false

ggplot() +
  geom_line(data = clim_pred, aes(x = as.Date(paste(year, 01, 01, sep = "-")), y = gdd_ings-meangdd/10, color = ID, group = ID), linewidth = 0.5) +
  scale_color_gradientn(colors = kippenberger) +
  theme_bw() +
  scale_x_date(expand = c(0,0), date_labels = "%Y", breaks =  as.Date(paste(seq(1980,2020,5), 01, 01, sep = "-"))) +
  theme(legend.position = 'none',
        panel.grid = element_blank()) +
  labs(x = "Year", y = "GDD accumulated in GS (anomaly, x10 degC)")

```