---
title: 'How do trees grow in Western North America:\newline A new modeling approach'
toc: false
number-sections: true
highlight: pygments
pdf-engine: lualatex
format:
  html:
    html-math-method: katex
    theme:
      - lux
      - custom.scss
    standalone: true
    embed-resources: true
    code-overflow: wrap
    linkcolor: "#B97C7C"
    fontsize: '11pt'
    code-block-border-left: "#8F2727"
    code-block-bg: "#f8f8f8"
    highlight-style: github
  pdf:
    keep-tex: true
    code-overflow: wrap
    monofont: 'Latin Modern Mono'
    monofontoptions:
      - Scale=0.5
format-links: false
knitr:
  opts_chunk:
    comment: ''
    cache.lazy: FALSE
  opts_knit:
    global.par: TRUE
geometry:
  - top=20mm
  - left=20mm
  - bottom=20mm
  - right=20mm
---

```{css, echo=FALSE}
.cell-output { border-left: #B97C7C 2px solid; border-radius: 0px;}
```


```{r}
#| echo: false
#| warning: false
#| message: false
wd <- "~/projects/climategrowthshifts/analysis/pnwandmore"
setwd(file.path(wd, 'model/shocks'))
util <- new.env()
source('mcmc_analysis_tools_rstan.R', local=util)
source('mcmc_visualization_tools.R', local=util)
```

# The motivations behind the project

::: {.callout-tip}
## The technical motivation

Bayesian magic! We fit all together the low frequency, decadal (to centennial) variations (the _detrending_) **and** the climatic variations (of interest). We let the model and the data speak, i.e. disentangle allometry (age/size trends in growth) from climate. This way, we avoid many arbitrary modeling decision (e.g. when one has to choose the frequency cut-off of a spline). And all the uncertainties get propagated to any decisions/predictions.
:::

\

::: {.callout-tip}
## The ecological motivation

Tree ring data offers offer an unique opportunity to learn how trees grow in a changing climate, through the complex interactions between external events and size-dependent developmental stages. While dendrochronology has mostly focus on past climate reconstructions, our approach do not try to standardize away the complexities of individual-level growth and species-specific variations. On the contrary, we believe that this complexity---if handle correctly by a model---can help us understand how tree growth varies within and **across species, ecosystems, and climates**. And _in fine_ our approach could improve our forecasts of forest dynamics, carbon sequestration, and thus climate change itself!

:::

# A quick summary of the model

Before diving into the results of our first inferences, let's describe the model! \
We can divide the model in two main parts:

- the mid- to long-term processes that are not directly related to climate, \newline such as **allometric constraints**, neighbour competition dynamics, ...

- the short-term variations that are related to external conditions, \newline mostly **climate**, but also pest outbreaks, ...

## Mid- to long-term processes

When we think about these low frequency variations, we mainly think about the age/size trends that affect tree growth, due to the natural size and shape of trees. We want this part of the model to be **flexible** enough to accommodates phase shifts, since tree growth is rarely perfectly smooth.\

```{r, fig.height=3.5, fig.width=8, fig.align="center"}
#| echo: false
#| fig-heigth: 2
#| fig-width: 8

par(family="serif", las=1, bty="l",
    cex.axis=0.8, cex.lab=0.85, cex.main=1,
    xaxs="i", yaxs="i", mar = c(5, 3, 3, 1))

gompertz <- function(dmax, beta, tau, t){
  return(dmax*exp(-log(2)*exp(-2/log(2)*beta/dmax*(t-tau))))
}

par(mfrow = c(1,2))

t <- seq(1900, 1984, 7)
dbh <- gompertz(dmax = 80, beta = 2, tau = 1940, t)
dbh[3] <- dbh[3] + 3; dbh[4] <- dbh[4] + 5; dbh[5] <- dbh[5] - 2;
dbh[8] <- dbh[8] - 6; dbh[9] <- dbh[9] - 12; dbh[10] <- dbh[10] - 8;
dbh[12] <- dbh[12] - 1; dbh[13] <- dbh[13] +3;
plot(dbh ~ t, type = 'p', axes = FALSE, frame = FALSE, xlab = '', ylab = '',
     pch = 20, col = util$c_mid, xlim = c(1900-1,2000+1), ylim = c(-2, 82))
title(xlab = 'Time', ylab = 'Diameter', line = 1, main = 'Observed diameter growth...')
axis(1, at = seq(1900-1, 2000+1, 1), tick = TRUE, labels = FALSE, lwd.ticks = 0)
axis(2, at = seq(-2, 82, 1), tick = TRUE, labels = FALSE, lwd.ticks = 0)


t2 <- seq(1900, 2000, 1)
dbh2 <- gompertz(dmax = 80, beta = 2, tau = 1940, t2)
plot(dbh ~ t, type = 'p', axes = FALSE, frame = FALSE, xlab = '', ylab = '',
     pch = 20, col = util$c_light, xlim = c(1900-1,2000+1), ylim = c(-2, 82), cex = 0.85)
lines(dbh2 ~ t2, type = 'l', lwd = 1.5, col = util$c_dark, lty = 2)
title(xlab = 'Time', ylab = 'Diameter', line = 1, main = '...is rarely smooth')
axis(1, at = seq(1900-1, 2000+1, 1), tick = TRUE, labels = FALSE, lwd.ticks = 0)
axis(2, at = seq(-2, 82, 1), tick = TRUE, labels = FALSE, lwd.ticks = 0)
```

We believe that these variations are "_not directly related to climate_", but not totally independent of the climate neither: a big storm could knock down surrounding trees and reduce competition for an individual.

We model these variations with a first **Gaussian process** (GP). The GP is characterized by a covariance function that captures the temporal correlation between ring widths that persists over one or several decades. The **length scale** $\mathbf{\rho}$ of this covariance function determined how quickly the correlations between ring widths decay. In the example below, ring width $x_2$ is more related to $x_1$ than $x_3$ is. \

```{r, fig.height=3.5, fig.width=8, fig.align="center"}
#| echo: false

expquad <- function(alpha, rho, deltat){
  return(alpha^2*exp(-1/2*(deltat/rho)^2))
}

deltat <- seq(0, 47, 1)
cov <- expquad(alpha = 1, rho = 15, deltat)
plot(cov ~ deltat, type = 'l', axes = FALSE, frame = FALSE, xlab = '', ylab = '',
    lwd = 1.5, col = util$c_light_highlight,, ylim = c(0, 1.1))
title(xlab = expression(paste(Delta, 'time (between two rings)')), ylab = 'Correlation', line = 1,
      main = 'A covariance function...')
axis(1, at = seq(0, 47, 1), tick = TRUE, labels = FALSE, lwd.ticks = 0)
axis(2, at = seq(0, 1.03, 0.01), tick = TRUE, labels = FALSE, lwd.ticks = 0)
points(x = c(0, 7, 20), y = expquad(alpha = 1, rho = 15, c(0,7,20)), pch = 20, col = 'white', xpd=NA, cex = 1.5)
points(x = c(0, 7, 20), y = expquad(alpha = 1, rho = 15, c(0,7,20)), pch = 20, col = util$c_dark, xpd=NA, cex = 1)
text(x = c(0, 7, 20), y = expquad(alpha = 1, rho = 15, c(0,7,20)) + 0.1, labels = c(expression(x[1]), expression(x[2]), expression(x[3])), col = util$c_dark, xpd=NA)

plot(cov ~ deltat, type = 'l', axes = FALSE, frame = FALSE, xlab = '', ylab = '',
    lwd = 1.5, col = util$c_light_highlight, ylim = c(0, 1.1))
title(main = expression(bold(paste('...mainly determined by a length scale ', rho))), line = 1)
segments(x0 = 15, x1 = 15, y0 = 0, y1 = 1*exp(-1/2), lty = 2, col = util$c_dark, lwd = 1.5)
segments(x0 = 0, x1 = 15, y0 = 1*exp(-1/2), y1 = 1*exp(-1/2), lty = 2, col = util$c_dark, lwd = 1.5)
# title(xlab = expression(paste(Delta, 'time (between two rings)')), ylab = 'Correlation', line = 1)
axis(1, at = c(0, 15, 47), tick = TRUE, labels = FALSE, lwd.ticks = 0)
axis(2, at = seq(0, 1.02, 0.01), tick = TRUE, labels = FALSE, lwd.ticks = 0)
text(x = 15, y = -0.07, labels = expression(rho), xpd=NA, col = util$c_dark)
text(x = 45, y = -0.07, labels = expression(3*rho), xpd=NA, col = util$c_dark)

```

The correlation across ring widths becomes negligeable around $3\rho$ years. \newpage

This mid- to long-term GP allows to capture, at the individual tree level, the low frequency variations that appear in the ring width measurements. \

```{r, fig.height=3.5, fig.width=6, fig.align="center"}
#| echo: false

example_samples <- readRDS("/home/victor/projects/climategrowthshifts/docs/reports/summary_forkevin/aux_files/example_samples.rds")
data <- readRDS(file = file.path('/home/victor/projects/climategrowthshifts/analysis/pnwandmore/model/improve_computation/output/model', 'data_05dec2025_PIPO_5stands.rds'))

par(family="serif", las=1, bty="l",
    cex.axis=0.8, cex.lab=0.85, cex.main=0.9,
    xaxs="i", yaxs="i", mar = c(5, 10, 3, 1))
t <- 20
idxs_t <- data$tree_start_idxs[t]:data$tree_end_idxs[t]

# names <- sapply(idxs_t,
#                 function(x) paste0('f[', x, ']'))
# util$plot_conn_pushforward_quantiles(example_samples, names, data$years[idxs_t],
#                                      xlab="", ylab='Ring width (log scale)', main = 'A long-term GP',
#                                      display_ylim=c(-0.5, 1.7), display_xlim = c(1979.5, 2010.5))
# points(data$years[idxs_t], ifelse(data$log_rw_obs[idxs_t] == log(1e-4), NA, data$log_rw_obs[idxs_t]), pch=16, cex=1, col="white")
# points(data$years[idxs_t], ifelse(data$log_rw_obs[idxs_t] == log(1e-4), NA, data$log_rw_obs[idxs_t]), pch=16, cex=0.6, col="black")
# abline(v =  c(2002,1996), lty = 3, col="darkgrey")
# 
# 
for(i in idxs_t){

  example_samples[[paste0('exp_', 'f[', i, ']')]] <- exp(example_samples[[paste0('f[', i, ']')]])

}

par(mfrow = c(1,1))
names <- sapply(idxs_t,
                function(x) paste0('exp_f[', x, ']'))
util$plot_conn_pushforward_quantiles(example_samples, names, data$years[idxs_t],
                                     xlab="", ylab='', main = 'Low frequency variations captured by the long-term GP',
                                     display_ylim=c(1, 5), display_xlim = c(1979.5, 2010.5))
title(ylab = 'Ring width (mm)', line = 2)
points(data$years[idxs_t], exp(ifelse(data$log_rw_obs[idxs_t] == log(1e-4), NA, data$log_rw_obs[idxs_t])), pch=16, cex=1, col="white")
points(data$years[idxs_t], exp(ifelse(data$log_rw_obs[idxs_t] == log(1e-4), NA, data$log_rw_obs[idxs_t])), pch=16, cex=0.6, col="black")
abline(v =  c(2002,1996), lty = 3, col="darkgrey")

```

Since we expect some difference across fast- or slow-growing species, we allow the length scale parameter to **vary across species**.

## Short-term variations

When we think about these low frequency variations, we mainly think about the yearly climatic conditions that explain short-term variations in growth. We first believed that these climatic effects were shared across conspecific trees within the same stand---in other words, that the trees from the same species on the same stand would have the same decrease/increase in growth because they experience similar climatic conditions. But we will see that it's a bit more complex than!

### Climatic predictors (stand-level)

This part of the model is quite simple. \
We selected three climatic predictors that are believed to have a strong impact on tree growth:

- growing degree days (GDD, between )  
- soil moisture (SM, ) -- which is mostly driven by winter precipitation  
- vapor pressure deficit (VPD, )  

::: {.callout-note}
We extracted climate from `WLDAS`,  which is available from 1980, at a X-m resolution. \
Since we plan to run our model from 1900 onward, we will also use `PRISM` (with simpler predictors, likely GDD and winter precipitation only).
:::

These 3 predictors are combined within a simple linear regression. We allow the effect of each predictor to **vary across species** (i.e. we estimate species-specific slopes $\beta$). \
This allows to identify different sensitivities between species, and differentiate between Gymnosperms (in blue) and Angiosperms (in green).


```{r, out.width="70%", fig.align="center"}
#| echo: false

knitr::include_graphics("/home/victor/projects/climategrowthshifts/analysis/pnwandmore/figures/newyork2025/model_estimates/climate_slopes_posterior_distributions.pdf")


```

In the left panel of this figure, each bar represents the posterior distribution of a species-specific slope. The right part shows the clade-level estimate (i.e. Gymnosperms or Angiosperms).

### Shared residual variations (stand-level)

The above predictors are not sufficient to capture the **shared variations** of growth across trees within the same stand. \
We added a short-term GP, initially with the goal to capture 3 to 5-year patterns. But the model consistently finds some yearly residual variation. In other words, the length scale of this GP is consistently around 1 year.

::: {.callout-note}
While the long-term GP was at the individual level (each tree gets its own individual GP, determined by a species-specific length scale), this short-term GP is shared across all trees in the same stand -- and we estimate only one length scale across all stands.
:::

### Shocks!

Show that it's on drought years, but some other years are similarly dry.