---
title: "Diameters and tree-rings"
format: 
  html:
    code-overflow: wrap
    page-layout: full
editor_options: 
  chunk_output_type: console
crossref:
  lst-title: "Stan Program"
filters:
   - include-code-files
---


```{r}

rm(list=ls())
wd <- "~/projects/climategrowthshifts/analysis/mountrainier"

```

## Tree-ring data from Ailene

```{r}

treerings <- read.csv(file.path(wd, "data", "treerings_ailene", "SouthSidecoresX.csv"), header=TRUE)
treerings$uniqueID <- paste(treerings$Tag, treerings$Core, sep = "_")                      
meas_cols <-  names(treerings)[grep("X", names(treerings))]
treerings <- reshape(treerings, dir = "long", 
                     varying = meas_cols, 
                     times = as.numeric(gsub("X", "", meas_cols)),
                     v.names = "rw", timevar = "year")
treerings <- treerings[c("Stand", "Species", "Tag", "Core", "uniqueID", "DBH", "RingCount", "year", "rw")]
treerings <- na.omit(treerings)
treerings <- treerings[order(treerings$year),]
treerings$cumrw <- ave(treerings$rw, treerings$uniqueID, FUN = cumsum)
unique(treerings$Stand)


```

## DBH data from PSP - until 2017

```{r}

tv01002 <- read.csv(file.path(wd, "data", "pnw_psp", "TV01002_v17.csv"))
mtr_data <- tv01002[tv01002$PSP_STUDYID == "MRRS",];rm(tv01002)
mtr_data <- na.omit(mtr_data[c("STANDID", "TREEID", "SPECIES", "DBH", "YEAR")]) # remove NA (death)
unique(mtr_data$STANDID)
unique(mtr_data$STANDID)[unique(mtr_data$STANDID) %in% unique(treerings$Stand)] # PARA?

```

## Quick look at both datasets

```{r}

# Number of trees, in common stands
merge(setNames(aggregate(data=mtr_data[which(mtr_data$STANDID %in% unique(treerings$Stand)),], TREEID ~ STANDID, function(x) length(unique(x))), c("stand", "n_trees_dbh")),
      setNames(aggregate(data=treerings[which(treerings$Stand %in% unique(mtr_data$STANDID)),], Tag ~ Stand, function(x) length(unique(x))), c("stand", "n_trees_rw")))

```

### Example from one plot, AV06

```{r}

dbh_data_av06 <- mtr_data[mtr_data$STANDID == "AV06",]
rw_data_av06 <- treerings[treerings$Stand == "AV06",]

```

```{r}
#| fig-height: 16
#| fig-width: 16

par(mfrow=c(10, 8) , mar = c(2,1.8,1.5,1.5))

for (t in unique(rw_data_av06$Tag)){
  
  pl.data <- rw_data_av06[rw_data_av06$Tag == t,]
  plot(pl.data$rw ~ pl.data$year, col = ifelse(pl.data$Core ==1, "grey50", "grey80"),
       pch = 19, cex = 0.5)

}

```

```{r}
#| fig-height: 16
#| fig-width: 16

par(mfrow=c(10, 8) , mar = c(2,1.8,1.5,1.5))

for (t in unique(rw_data_av06$Tag)){
  
  pl.data <- rw_data_av06[rw_data_av06$Tag == t,]
  plot(pl.data$cumrw ~ pl.data$year, col = ifelse(pl.data$Core ==1, "grey50", "grey80"),
       pch = 19, cex = 0.3)

}

```

## Updated DBH data from PSP - received on Feb. 2025

```{r}

mtr_2024 <- data.frame(readxl::read_xlsx(file.path(wd, "data", "pnw_psp", "MRRS_2024", "trees_MRRS_2024.xlsx")))

```

Two plots are also missing in the PSP dataset:

```{r}

unique(mtr_2024$Stand)[!(unique(mtr_2024$Stand) %in% unique(mtr_data$STANDID))]

mtr_2024_sub <- mtr_2024[mtr_2024$Stand %in% unique(mtr_data$STANDID), ]

```

If we look only at the common plots, 2024 dataset has `{r} length(unique(mtr_2024_sub$TreeID))` trees, and PSP dataset has `{r} length(unique(mtr_data[mtr_data$YEAR == 2017, "TREEID"]))` trees living in 2017 (in the `{r} length(unique(mtr_2024_sub$Stand))` shared plots).\
Some tree IDs are duplicated in 2024 dataset.

```{r}

dup_ids <- mtr_2024_sub[duplicated(mtr_2024_sub$TreeID) & !is.na(mtr_2024_sub$TreeID), "TreeID"]

```

So,  `{r} length(unique(mtr_data[mtr_data$YEAR == 2017, "TREEID"]))- length(unique(mtr_2024_sub$TreeID))` trees are missing. There are `{r} nrow(mtr_2024_sub[is.na(mtr_2024_sub$TreeID), ])` trees without IDs in the 2024 dataset, and `{r} length(dup_ids)` with duplicated IDs.
Let's look at those.

### Duplicated trees

```{r}

for(i in dup_ids){
  
  
  
}

```


### Unidentified trees

```{r}

dbhs_noid <- as.numeric(mtr_2024_sub[is.na(mtr_2024_sub$TreeID), "Dbh"])

par(mfrow=c(1, 1))
hist(dbhs_noid, main = "DBHs of trees with missing IDs", xlab = "DBH", breaks = seq(min(dbhs_noid), max(dbhs_noid), l = 30))

```

There are mostly small trees, which have just reached DBH threshold of 5 cm---but not only.\
Coincidentally, trees with missing IDs are the only ones where columns related to canopy (`Canopy_class`, `Crown_ratio`, `Azimuth`...) are not ``NA`.\

Let's do a quick check on unidentified trees with DBH > 20cm, just to start (`{r} nrow(mtr_2024_sub[is.na(mtr_2024_sub$TreeID) & mtr_2024_sub$Dbh >= 20, ])` trees).

```{r}

bigtrees_noid <- mtr_2024_sub[is.na(mtr_2024_sub$TreeID) & mtr_2024_sub$Dbh > 20, ]

grepl("Fused|Shares", paste(bigtrees_noid$Comment_1, bigtrees_noid$Comment_2, bigtrees_noid$Comment_3, bigtrees_noid$Comment_4))

```

Some of these trees are fused, or shared a common base, with another tree---which could explain why they don't have an ID?\
Let's look at the others.

```{r}

bigtrees_noid_noshare <- bigtrees_noid[!grepl("Fused|Shares", paste(bigtrees_noid$Comment_1, bigtrees_noid$Comment_2, bigtrees_noid$Comment_3, bigtrees_noid$Comment_4)),]

```

`TREEID` variable is made with `STANDID` and `PLOTNUMBER`.\
For example, a tree on stand  `AR19` and plot `5` would have an ID beginning by `AR190005`.\
It might help us to find some missing trees?

```{r}

wheretolook <- paste0(bigtrees_noid_noshare$Stand, ifelse(bigtrees_noid_noshare$Plot < 10, paste0("000", bigtrees_noid_noshare$Plot), paste0("00", bigtrees_noid_noshare$Plot)))

for(i in 1:length(wheretolook)){
  
  dbh_threshold <- bigtrees_noid_noshare[i, "Dbh"]-2  # -2, not to be too restrictive
  sp <- bigtrees_noid_noshare[i, "Species"]
  potential_trees <- unique(mtr_data[grepl(wheretolook[i], mtr_data$TREEID) & 
                                       mtr_data$YEAR %in% c(2017) & 
                                       mtr_data$DBH < dbh_threshold &
                                       mtr_data$SPECIES %in% sp, ])
  
  already_present <- mtr_2024_sub[mtr_2024_sub$TreeID %in% na.omit(potential_trees$TREEID), ]
  
  cat(paste0(nrow(potential_trees)-nrow(already_present), " "))
  
}

```

We get only zeros, meaning that all potential trees in the PSP dataset (before 2024) that could correspond to trees with missing IDs are already in the 2024 dataset!
We could try to do the same thing, with trees with DBH between 10 and 20 cm.

```{r}

bigtrees_noid <- mtr_2024_sub[is.na(mtr_2024_sub$TreeID) & mtr_2024_sub$Dbh <= 20 &  mtr_2024_sub$Dbh >= 10, ]
bigtrees_noid_noshare <- bigtrees_noid[!grepl("Fused|Shares", paste(bigtrees_noid$Comment_1, bigtrees_noid$Comment_2, bigtrees_noid$Comment_3, bigtrees_noid$Comment_4)),]
wheretolook <- paste0(bigtrees_noid_noshare$Stand, ifelse(bigtrees_noid_noshare$Plot < 10, paste0("000", bigtrees_noid_noshare$Plot), paste0("00", bigtrees_noid_noshare$Plot)))

for(i in 1:length(wheretolook)){
  print(i)
  dbh_threshold <- bigtrees_noid_noshare[i, "Dbh"]-2  # -2, not to be too restrictive
  sp <- bigtrees_noid_noshare[i, "Species"]
  potential_trees <- unique(mtr_data[grepl(wheretolook[i], mtr_data$TREEID) & 
                                       mtr_data$YEAR %in% c(2013,2017) & 
                                       mtr_data$DBH < dbh_threshold &
                                       mtr_data$SPECIES %in% sp, ])
  
  already_present <- mtr_2024_sub[mtr_2024_sub$TreeID %in% na.omit(potential_trees$TREEID), c("Stand", "TreeID", 'Dbh')]
  
  cat(paste0(nrow(potential_trees)-nrow(already_present), " "))
  
}

```

