---
title: "Diameters and tree-rings"
format: 
  html:
    code-overflow: wrap
    page-layout: full
editor_options: 
  chunk_output_type: console
crossref:
  lst-title: "Stan Program"
filters:
   - include-code-files
---


```{r}

rm(list=ls())
wd <- "~/projects/climategrowthshifts/analysis/mountrainier"

```

## Tree-ring data from Ailene

```{r}

treerings <- read.csv(file.path(wd, "data", "treerings_ailene", "SouthSidecoresX.csv"), header=TRUE)
treerings$uniqueID <- paste(treerings$Tag, treerings$Core, sep = "_")                      
meas_cols <-  names(treerings)[grep("X", names(treerings))]
treerings <- reshape(treerings, dir = "long", 
                     varying = meas_cols, 
                     times = as.numeric(gsub("X", "", meas_cols)),
                     v.names = "rw", timevar = "year")
treerings <- treerings[c("Stand", "Species", "Tag", "Core", "uniqueID", "DBH", "RingCount", "year", "rw")]
treerings <- na.omit(treerings)
treerings <- treerings[order(treerings$year),]
treerings$cumrw <- ave(treerings$rw, treerings$uniqueID, FUN = cumsum)
unique(treerings$Stand)


```

## DBH data from PSP - until 2017

```{r}

tv01002 <- read.csv(file.path(wd, "data", "pnw_psp", "TV01002_v17.csv"))
mtr_data <- tv01002[tv01002$PSP_STUDYID == "MRRS",];rm(tv01002)
mtr_data <- mtr_data[!is.na(mtr_data$DBH),]# remove NA (death)
unique(mtr_data$STANDID)
unique(mtr_data$STANDID)[unique(mtr_data$STANDID) %in% unique(treerings$Stand)] # PARA?

```

## Quick look at both datasets

```{r}

# Number of trees, in common stands
merge(setNames(aggregate(data=mtr_data[which(mtr_data$STANDID %in% unique(treerings$Stand)),], TREEID ~ STANDID, function(x) length(unique(x))), c("stand", "n_trees_dbh")),
      setNames(aggregate(data=treerings[which(treerings$Stand %in% unique(mtr_data$STANDID)),], Tag ~ Stand, function(x) length(unique(x))), c("stand", "n_trees_rw")))

```

### Example from one plot, AV06

```{r}

dbh_data_av06 <- mtr_data[mtr_data$STANDID == "AV06",]
rw_data_av06 <- treerings[treerings$Stand == "AV06",]

```

```{r}
#| fig-height: 16
#| fig-width: 16

par(mfrow=c(10, 8) , mar = c(2,1.8,1.5,1.5))

for (t in unique(rw_data_av06$Tag)){
  
  pl.data <- rw_data_av06[rw_data_av06$Tag == t,]
  plot(pl.data$rw ~ pl.data$year, col = ifelse(pl.data$Core ==1, "grey50", "grey80"),
       pch = 19, cex = 0.5)

}

```

```{r}
#| fig-height: 16
#| fig-width: 16

par(mfrow=c(10, 8) , mar = c(2,1.8,1.5,1.5))

for (t in unique(rw_data_av06$Tag)){
  
  pl.data <- rw_data_av06[rw_data_av06$Tag == t,]
  plot(pl.data$cumrw ~ pl.data$year, col = ifelse(pl.data$Core ==1, "grey50", "grey80"),
       pch = 19, cex = 0.3)

}

```

## Updated DBH data from PSP - received on Feb. 2025

```{r}

mtr_2024 <- data.frame(readxl::read_xlsx(file.path(wd, "data", "pnw_psp", "MRRS_2024", "trees_MRRS_2024.xlsx")))

```

Two plots are also missing in the PSP dataset:

```{r}

unique(mtr_2024$Stand)[!(unique(mtr_2024$Stand) %in% unique(mtr_data$STANDID))]

mtr_2024_sub <- mtr_2024[mtr_2024$Stand %in% unique(mtr_data$STANDID), ]

```

If we look only at the common plots, 2024 dataset has `{r} length(unique(mtr_2024_sub$TreeID))` trees, and PSP dataset has `{r} length(unique(mtr_data[mtr_data$YEAR == 2017, "TREEID"]))` trees living in 2017 (in the `{r} length(unique(mtr_2024_sub$Stand))` shared plots).\
Some tree IDs are duplicated in 2024 dataset.

```{r}

dup_ids <- mtr_2024_sub[duplicated(mtr_2024_sub$TreeID) & !is.na(mtr_2024_sub$TreeID), "TreeID"]

```

So,  `{r} length(unique(mtr_data[mtr_data$YEAR == 2017, "TREEID"]))- length(unique(mtr_2024_sub$TreeID))` trees are missing. There are `{r} nrow(mtr_2024_sub[is.na(mtr_2024_sub$TreeID), ])` trees without IDs in the 2024 dataset, and `{r} length(dup_ids)` with duplicated IDs.
Let's look at those.

### Duplicated trees

```{r}

mtr_2024_dup <- mtr_2024_sub[mtr_2024_sub$TreeID %in% dup_ids, ]

for(i in dup_ids){
  
  cat(paste0(nrow(mtr_2024_dup[mtr_2024_dup$TreeID %in% i, ]), " "))

}

```

There is never more than two records (yeah!).

Some duplicated trees are just pure duplicates (i.e. exactly the same data, including DBH).\
It's quite straightforward to remove them, without any risk.

```{r}

mtr_2024_dup <- unique(mtr_2024_dup)
dup_ids <- mtr_2024_dup[duplicated(mtr_2024_dup$TreeID), "TreeID"]

```

Then, we have only `{r} length(dup_ids)` trees with duplicated IDs and different data.

```{r}

mtr_2024_dup <- mtr_2024_sub[mtr_2024_sub$TreeID %in% dup_ids, ]
save <- data.frame(diff = NA)

n <- 1
for(i in dup_ids){
  
  d <- mtr_2024_dup[mtr_2024_dup$TreeID %in% i, ]
  cat(paste0(round(d[1,"Dbh"] - d[2,"Dbh"],3), " "))

}

```

Some DBHs are really similar, some are reaaaally different.\
Example, this striking example: `{r} dup_ids[15]`.

```{r}

i <-  dup_ids[15]
mtr_2024_dup[mtr_2024_dup$TreeID %in% i,c("TreeID", "Status", "Dbh", paste0("Comment_", 1:4))]

```

Based on PSP data, the DBH of 43.1 seems more likely.

```{r}

mtr_data[mtr_data$TREEID %in% i, c("YEAR", "DBH")]

```

Let's now look at `{r} dup_ids[14]`

```{r}

i <-  dup_ids[14]
mtr_2024_dup[mtr_2024_dup$TreeID %in% i,c("TreeID", "Status", "Dbh", paste0("Comment_", 1:4))]

```

Tree is both recorded as missing or not found (`Status = 9`), and also as living (and measured, `Status = 1`)...\

I guess the best we can do now is to clean these `{r} length(dup_ids)` trees manually (if possible?!).

```{r}

i <-  dup_ids[1]
mtr_2024_dup[mtr_2024_dup$TreeID %in% i,c("TreeID", "Status", "Dbh", paste0("Comment_", 1:4))] # DBH 55 or 53.7

i <-  dup_ids[2]
mtr_2024_dup[mtr_2024_dup$TreeID %in% i,c("TreeID", "Status", "Dbh", paste0("Comment_", 1:4))] # DBH 44.5 or 44.6

i <-  dup_ids[3]
mtr_2024_dup[mtr_2024_dup$TreeID %in% i,c("TreeID", "Status", "Dbh", paste0("Comment_", 1:4))] # same DBH, one without any comment
mtr_2024_sub <- mtr_2024_sub[-which(mtr_2024_sub$TreeID %in% dup_ids[3] & is.na(mtr_2024_sub$Comment_1)), ] # remove it

i <-  dup_ids[4]
mtr_2024_dup[mtr_2024_dup$TreeID %in% i,c("TreeID", "Status", "Dbh", paste0("Comment_", 1:4))] # DBH 43.5 or 43.8

i <-  dup_ids[5]
mtr_2024_dup[mtr_2024_dup$TreeID %in% i,c("TreeID", "Status", "Dbh", paste0("Comment_", 1:4))] # DBH 14.3 or 14.2 - one with less comments
mtr_2024_sub <- mtr_2024_sub[-which(mtr_2024_sub$TreeID %in% dup_ids[5] & is.na(mtr_2024_sub$Comment_2)), ] # remove it

i <-  dup_ids[6]
mtr_2024_dup[mtr_2024_dup$TreeID %in% i,c("TreeID", "Status", "Dbh", paste0("Comment_", 1:4))] # same DBH, one without any comment
mtr_2024_sub <- mtr_2024_sub[-which(mtr_2024_sub$TreeID %in% dup_ids[6] & is.na(mtr_2024_sub$Comment_1)), ] # remove it

i <-  dup_ids[7]
mtr_2024_dup[mtr_2024_dup$TreeID %in% i,c("TreeID", "Status", "Dbh", paste0("Comment_", 1:4))] # DBH 96.4 or 81.1, big diff.
mtr_data[mtr_data$TREEID %in% i, c("YEAR", "DBH")] # DBH in 2017 = 82.1, so neither 96.4 nor 81.1 are likely

i <-  dup_ids[8]
mtr_2024_dup[mtr_2024_dup$TreeID %in% i,c("TreeID", "Status", "Dbh", paste0("Comment_", 1:4), "Lean_deg")] # same DBH, Lean deg 0 or 5
mtr_2024_sub <- mtr_2024_sub[-which(mtr_2024_sub$TreeID %in% dup_ids[8] & mtr_2024_sub$Lean_deg == 0), ] # remove 0

i <-  dup_ids[9]
mtr_2024_dup[mtr_2024_dup$TreeID %in% i,c("TreeID", "Status", "Dbh", paste0("Comment_", 1:4))] # One with DBH NA
mtr_2024_sub <- mtr_2024_sub[-which(mtr_2024_sub$TreeID %in% dup_ids[9] & is.na(mtr_2024_sub$Dbh)), ] # remove it

i <-  dup_ids[10]
mtr_2024_dup[mtr_2024_dup$TreeID %in% i,c("TreeID", "Status", "Dbh", paste0("Comment_", 1:4))] # DBH 7.1 or 7.3

i <-  dup_ids[11]
mtr_2024_dup[mtr_2024_dup$TreeID %in% i,c("TreeID", "Status", "Dbh", paste0("Comment_", 1:4))] # DBH 17.6 or 16 + one without comment...
mtr_data[mtr_data$TREEID %in% i, c("YEAR", "DBH")] # DBH in 2017 = 15.4, in 2012 = 15.3 (DBH = 16 more likely?)

i <-  dup_ids[12]
mtr_2024_dup[mtr_2024_dup$TreeID %in% i,c("TreeID", "Status", "Dbh", paste0("Comment_", 1:4))] # same DBH, one with less comments
mtr_2024_sub <- mtr_2024_sub[-which(mtr_2024_sub$TreeID %in% dup_ids[12] & is.na(mtr_2024_sub$Comment_2)), ] # remove it

i <-  dup_ids[13]
mtr_2024_dup[mtr_2024_dup$TreeID %in% i,c("TreeID", "Status", "Dbh", paste0("Comment_", 1:4))] # DBH 34.8 or 34.5 + different comments

i <-  dup_ids[14]
mtr_2024_dup[mtr_2024_dup$TreeID %in% i,c("TreeID", "Status", "Dbh", paste0("Comment_", 1:4))] # One with DBH NA
mtr_2024_sub <- mtr_2024_sub[-which(mtr_2024_sub$TreeID %in% dup_ids[14] & is.na(mtr_2024_sub$Dbh)), ] # remove it

i <-  dup_ids[15]
mtr_2024_dup[mtr_2024_dup$TreeID %in% i,c("TreeID", "Status", "Dbh", paste0("Comment_", 1:4))] # DBH 68.9 or 43.1, big diff.
mtr_data[mtr_data$TREEID %in% i, c("YEAR", "DBH")] # DBH in 2017 = 41.6, so DBH = 43.1 more likely
mtr_2024_sub <- mtr_2024_sub[-which(mtr_2024_sub$TreeID %in% dup_ids[15] & mtr_2024_sub$Dbh == 43.1), ] # remove DBH = 68.9

i <-  dup_ids[16]
mtr_2024_dup[mtr_2024_dup$TreeID %in% i,c("TreeID", "Status", "Dbh", paste0("Comment_", 1:4))] # DBH 14.0 and 14.1 + different comments

```

### Unidentified trees

```{r}

dbhs_noid <- as.numeric(mtr_2024_sub[is.na(mtr_2024_sub$TreeID), "Dbh"])

par(mfrow=c(1, 1))
hist(dbhs_noid, main = "DBHs of trees with missing IDs", xlab = "DBH", breaks = seq(min(dbhs_noid), max(dbhs_noid), l = 30))

```

There are mostly small trees, which have just reached DBH threshold of 5 cm---but not only.\
There are `{r} nrow(mtr_2024_sub[is.na(mtr_2024_sub$TreeID) & mtr_2024_sub$Dbh >= 20, ])` trees with a DBh > 20 cm!
Ingrowth trees should be coded with `TREE_STATUS =2` in the database.

```{r}

dbhs_noid <- as.numeric(mtr_2024_sub[is.na(mtr_2024_sub$TreeID) & mtr_2024_sub$Status != 2, "Dbh"])


hist(dbhs_noid, main = "DBHs of trees with missing IDs", xlab = "DBH", breaks = seq(min(dbhs_noid), max(dbhs_noid), l = 30))

```

Removing ingrowth trees let only `{r} length(dbhs_noid)` trees (so some ingrowth trees have DBH > 20! Waou).

