---
title: "Explore gradient issues"
format: pdf
editor: visual
editor_options: 
  chunk_output_type: console
---

```{r}
suppressPackageStartupMessages(library(doFuture))
library(ggplot2)
suppressPackageStartupMessages(library(rstan))
options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)

wd <- "~/projects/climategrowthshifts/analysis/mountrainier"

# load data for which we got init failure in first_handling.R 
mdl.data <- readRDS(file.path(wd, 'output', 'saved_mdldata_initfailure.rds'))
```

## Brute-force approach

We define a function to set initial values for sampling. We draw uniformly from the interval (-2,2), the default strategy in Stan (on the unconstrained scale). For parameters that are bounded below at 0 (all except `delta_t_half`), we apply the exponential.

```{r}
initfun <- function() {
  list(d0 = runif(mdl.data$N_trees,exp(-2),exp(2)), 
       delta_d = runif(mdl.data$N_trees,exp(-2),exp(2)),
       beta = runif(mdl.data$N_trees,exp(-2),exp(2)), 
       delta_t_half = runif(mdl.data$N_trees,-2,2),
       sigma = runif(1,exp(-2),exp(2)))
}

```

Then we can run 20000 different Stan sampling (luckily, in parallel) to explore gradient failure.

```{r}
#| output: false

gompertz <- stan_model(file.path(wd, "stan/heterogeneous_gompertz.stan"))
plan(multisession, workers = 20)
init_fail <- foreach(i = 1:20000, .combine = rbind, 
                      .options.future = list(seed = TRUE)) %dofuture%{
  set.seed(sample.int(.Machine$integer.max, 1))
  
  inits <- initfun()
  fit <-  sampling(gompertz, mdl.data, 
                  chains = 1, cores = 1,
                  iter = 1, warmup = 0,
                  init = list(inits))
  
  data.frame(failure = ifelse(is.na(nrow(fit)), 'Failure', 'Success'),
             min.d0 = log(min(inits$d0)),
             min.delta_d = log(min(inits$delta_d)),
             min.beta = log(min(inits$beta)),
             min.delta_t_half = min(inits$delta_t_half),
             max.d0 = log(max(inits$d0)),
             max.delta_d = log(max(inits$delta_d)),
             max.beta = log(max(inits$beta)),
             max.delta_t_half = max(inits$delta_t_half),
             sigma = log(inits$sigma))

}
plan(sequential)

perc_fail <- nrow(init_fail[init_fail$failure == "Failure",])/nrow(init_fail)*100
```

Sampling fails in `{r} perc_fail`% of cases.\
The only parameter where there is an obvious difference between "Failure" and "Success" is delta_d (the difference between minimum and maximum DBH).

```{r}
#| echo: false

variables <- c("min.d0", "min.delta_d", "min.beta", "min.delta_t_half", "max.d0", "max.delta_d", "max.beta", "max.delta_t_half", "sigma")
init_fail_long <- reshape(init_fail, dir = "long",
                          v.names = "value", timevar = "var",
                          varying = variables, 
                          times = variables)

ggplot(data = init_fail_long[init_fail_long$var %in% c("min.d0", "min.delta_d", "min.beta", "min.delta_t_half"),]) +
  facet_wrap(failure ~ var, scales = 'free_y', nrow = 2) +
  geom_histogram(aes(x = value), alpha = 0.5, bins = 100) +
  theme_bw() +
  labs(y = "Count", x = "Min. init. parameter value")

ggplot(data = init_fail_long[init_fail_long$var %in% c("max.d0", "max.delta_d", "max.beta", "max.delta_t_half"),]) +
  facet_wrap(failure ~ var, scales = 'free_y', nrow = 2) +
  geom_histogram(aes(x = value), alpha = 0.5, bins = 100) +
  theme_bw() +
  labs(y = "Count", x = "Max. init. parameter value")

ggplot(data = init_fail_long[init_fail_long$var %in% c("sigma"),]) +
  facet_wrap(failure ~ var, scales = 'free_y', nrow = 2) +
  geom_histogram(aes(x = value), alpha = 0.5, bins = 100) +
  theme_bw() +
  labs(y = "Count", x = "Init. parameter value")

```

### Looking at delta_d

Taking the Gompertz function (reparameterized): $$
G(t)=d_{max}*exp(-log(2)*exp(-\frac{2}{log(2)}*\frac{\beta}{d_{max}}*(t-\tau)) 
$$ If we write: $$
D = log(2) * exp(-\frac{2}{log(2)}*\frac{\beta}{d_{max}}*(t-\tau))
$$ We have: $$
G(t) = d_{max}*exp(-D)
$$ Then, the partial derivative with respect to dmax (`delta_d`) should be something like: $$
\frac{\partial G}{\partial d_{max}} = exp(-D) (1-\frac{2}{log(2)} \frac{\beta}{{d_{max}}} (t-\tau)D)
$$

```{r}
f = expression(d_max * exp(-log(2) * exp( -   (2 / log(2)) * (beta / d_max) * 
                                            (t - (2000 + delta_t_half)))))
df_ddmax <- D(f,'d_max')

d0 <- 3
beta <- 1 # growth rate
delta_t_half <- -10
d_max <- 10

t <- 1980:2010
y <- eval(f)
plot(y ~ t)

d_max <- 10

t <- 1980
d_max <- seq(7,10, 0.0001)
y <- eval(df_ddmax)
plot(y ~ d_max)

```

Looking at $D = log(2) * exp(-\frac{2}{log(2)}*\frac{\beta}{d_{max}}*(t-\tau))$:

```{r}
f = expression(log(2) * exp(-(2 / log(2)) * (beta / x) * 
                              (t - (2000 + delta_t_half)))) 
x <- seq(0,0.1,0.00001)
y <- eval(f)
plot(y ~ x)

beta <- 5
f = expression(log(2) * exp(-(2 / log(2)) * (beta / x) * 
                              (t - (2000 + delta_t_half)))) 
x <- seq(0,0.5,0.00001)
y <- eval(f)
plot(y ~ x)

```

$D$ quickly become very high for some values of $d_{max}$, at least when $(t - \tau) < 0$.
In my example, $\tau = 2000-10$, we then get $-(2 / log(2)) * (t - \tau) \simeq 28.85$, which quickly drive the exponential towards infinity. 



