---
title: "Navigating in a shocking world"
pdf-engine: lualatex
format:
  pdf:
    keep-tex: true
    fig-width: 7
    fig-height: 3
    code-overflow: wrap
monofont: 'Latin Modern Mono'
monofontoptions: 'Scale=0.75'
knitr:
  opts_chunk:
    comment: ''
    cache.lazy: FALSE
  opts_knit:
    global.par: TRUE
---


It's time to expand the dataset I'm using. \
I look at all the *Pinus ponderosa* in Arizona, i.e. 914 trees on 20 stands (from 1980 onwards). \
Look at the end of the document to see the Stan model

```{r}
#| echo: false
#| warning: false
#| message: false
wd <- "~/projects/climategrowthshifts/analysis/pnwandmore"
setwd(file.path(wd, 'model'))
util <- new.env()
source('mcmc_analysis_tools_rstan.R', local=util)
source('mcmc_visualization_tools.R', local=util)
source('functions/custom_functions.R', local=util)
setwd(wd)
library(rstan)
options(mc.cores = parallel::detectCores())

chunk_hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- chunk_hook(x, options)
  paste0("\\linespread{0.5}\n", x, "\n\n\\linespread{2}")
})

par(family="serif", las=1, bty="l",
    cex.axis=0.8, cex.lab=0.85, cex.main=1,
    xaxs="i", yaxs="i", mar = c(5, 5, 3, 1))
```

## The data

```{r}
data <- readRDS(file = file.path(wd, 'output/model', 'data_15nov2025_azPIPO.rds'))

par(mfrow = c(1,2))
hist(data$log_rw_obs, col = util$c_mid, border = 'white', main ='',
     xlab = "Log ringwidth")
hist(data$log_rw_obs, col = util$c_mid, border = 'white', main ='',
     xlab = "Log ringwidth",
     ylim = c(0,500))

```

As a first guess, -4 seems like a good enough limit to define the (non-)centered configuration:

```{r}
data$w_sck <- ifelse(data$log_rw_obs < -4, 1, 0)
```

## A first fit

```{r}

run <- FALSE
if(run){
  fit <- stan(file=file.path(wd, 'model', 'stan', 
                             'model8_with3predictors_nopooling_standphi_nonconcordantshock.stan'),
            data=data, seed=5838293,
            chains = 4,
            warmup=1000, iter=2024, refresh=10)
  saveRDS(fit, file.path(wd, 'model/shocks/output/azPIPO', 'fit.rds'))
}else{
  fit <- readRDS(file.path(wd, 'model/shocks/output/azPIPO', 'fit.rds'))
}

```

Let's look at the diagnostics!

```{r}
diagnostics1 <- util$extract_hmc_diagnostics(fit)
util$check_all_hmc_diagnostics(diagnostics1)
```

There are many signs that the posterior computation might be inaccurate.
As a first step let's look at $\sigma$ and $\tau^\text{inner}_\text{shock}$, because we expect a degeneracy there.

```{r}
#| fig-width: 5
#| fig-height: 5
  
samples1 <- util$extract_expectand_vals(fit)
base_samples <- util$filter_expectands(samples1, c('sigma', 'inner_tau_sck'))
util$check_all_expectand_diagnostics(base_samples)

util$plot_pairs_by_chain(log(samples1[['sigma']]), 'log(sigma)', 
                         log(samples1[['inner_tau_sck']]), 'log(inner_tau_sck)')
```

HMC has trouble exploring the geometry here. Maybe because of some obstruction in exploring $\tau^\text{inner}_\text{shock}$ because of our crude initial (non-)centered configuration? \

### Shock geometry

Let's look at the individual shocks.

```{r}
name_shocks <- paste("delta_tilde_sck[", 1:data$N, "]", sep='')
median_shocks <- sapply(name_shocks, function(x) util$ensemble_mcmc_quantile_est(samples1[[x]], 0.5))


par(mfrow = c(1,2))
hist(median_shocks, col = util$c_mid, border = 'white', main ='',
     xlab = expression(delta[shock]), breaks = 30)
hist(median_shocks, col = util$c_mid, border = 'white', main ='',
     xlab = expression(delta[shock]), breaks = 30,
     ylim = c(0,100))
```

We can check for some evidence of big shocks that were not centered... For example, the first biggest negative shock that was not centered:
```{r}
#| fig-width: 5
#| fig-height: 5

last10mins <- sort(median_shocks[data$w_sck == 0])[1:10]

name_x <- sub("\\..*$", "", names(last10mins[1]))
util$plot_pairs_by_chain(samples1[[name_x]], name_x, 
                         log(samples1[['inner_tau_sck']]), 'log(inner_tau_sck)')

```

Or the second biggest negative shock that was not centered:
```{r}
#| fig-width: 5
#| fig-height: 5
name_x <- sub("\\..*$", "", names(last10mins[2]))
util$plot_pairs_by_chain(samples1[[name_x]], name_x, 
                         log(samples1[['inner_tau_sck']]), 'log(inner_tau_sck)')

```
And for a big positive shock:
```{r}
#| fig-width: 5
#| fig-height: 5

first10maxs <- sort(median_shocks[data$w_sck == 0], decreasing = TRUE)[1:10]

name_x <- sub("\\..*$", "", names(first10maxs[1]))
util$plot_pairs_by_chain(samples1[[name_x]], name_x, 
                         log(samples1[['inner_tau_sck']]), 'log(inner_tau_sck)')

```

However, we don't see any evidence of non-shocks that we should not have centered:
```{r}
#| fig-width: 5
#| fig-height: 5

absmins <- sort(abs(median_shocks[data$w_sck == 1]))[1:10]

name_x <- sub("\\..*$", "", names(absmins[1]))
util$plot_pairs_by_chain(samples1[[name_x]], name_x, 
                         log(samples1[['inner_tau_sck']]), 'log(inner_tau_sck)')

```
```{r}
#| fig-width: 5
#| fig-height: 5

absmins <- sort(abs(median_shocks[data$w_sck == 1]))[1:10]

name_x <- sub("\\..*$", "", names(absmins[2]))
util$plot_pairs_by_chain(samples1[[name_x]], name_x, 
                         log(samples1[['inner_tau_sck']]), 'log(inner_tau_sck)')

```

### Comparing shocks across stands

A first approach to identify if our shock model is capturing lack of growth as we want is to look at the stand-level GPs. The dashed line corresponds to 2002, because we expect it to be a shock year.
```{r}
#| fig-width: 7
#| fig-height: 9

par(mfrow = c(4,3))
for(s in 1:data$N_stands){
  names <- sapply(1:data$N_all_years,
                  function(y) paste0('f_sh[', s, ',', y, ']'))
  util$plot_realizations(samples1, names, plot_xs = data$all_years[1:data$N_all_years], N_plots = 50,
                         main = paste0('Stand ', data$uniq_stand_ids[s]), ylab = 'f_sh',
                         display_xlim = data$all_years[c(1,data$N_all_years)], display_ylim = c(-2,2))
  abline(v = 2002, lty = 2)
}
```

The stand-level GPs are still all very wiggly, but in particular we see evidence for some shocks in S11 and S15. \
Let's look more closely at individual shocks for trees in S11.

```{r}
#| fig-width: 7
#| fig-height: 9

trees_S11 <- which(data$stand_idxs == which(data$uniq_stand_ids == 'S11'))
par(mfrow = c(5,4))
for(t in trees_S11){
  idxs_t <- data$tree_start_idxs[t]:data$tree_end_idxs[t]
  names <- sapply(idxs_t,
                  function(x) paste0('delta_sck[', x, ']'))
  util$plot_realizations(samples1, names, plot_xs = data$years[idxs_t], N_plots = 100,
                         ylab = expression(delta[shocks]),
                         display_xlim = data$all_years[c(1,length(data$all_years))], 
                         display_ylim = c(-10,10),
                         main = paste0('Tree ', t))
}

```

Mostly flat lines, i.e. our shock model does not capture anything. \
If we look at another stand (S1), we see evidence for a more coherent behavior with what we expect

```{r}
#| fig-width: 7
#| fig-height: 9

trees_S1 <- which(data$stand_idxs == which(data$uniq_stand_ids == 'S1'))
par(mfrow = c(5,4))
for(t in trees_S1){
  idxs_t <- data$tree_start_idxs[t]:data$tree_end_idxs[t]
  names <- sapply(idxs_t,
                  function(x) paste0('delta_sck[', x, ']'))
  util$plot_realizations(samples1, names, plot_xs = data$years[idxs_t], N_plots = 100,
                         ylab = expression(delta[shocks]),
                         display_xlim = data$all_years[c(1,length(data$all_years))], 
                         display_ylim = c(-10,10),
                         main = paste0('Tree ', t))
}

```

These two stands, S1 and S11, get a very different probability of shocks at the stand level.

```{r}
#| fig-width: 4
#| fig-height: 3

par(mfrow = c(1,1))
util$plot_expectand_pushforward(samples1[[paste0('phi_sck[',which(data$uniq_stand_ids == 'S1'),']')]], 
                                50, flim = c(0,1), ylim = c(0,30),
                                display_name=expression(phi[shock]))
text("S1", x = 0.2, y = 7, col = util$c_dark)
util$plot_expectand_pushforward(samples1[[paste0('phi_sck[',which(data$uniq_stand_ids == 'S11'),']')]], 
                                50, flim = c(0,1), ylim = c(0,30),
                                display_name=expression(phi[shock]), add = TRUE, 
                                col = util$c_light_highlight)
text("S11", x = 0.05, y = 20, col = util$c_light_highlight)

```

## The model


```{r}
# we skip the function block
# many stuff commented out because I'm looking at only one species...
writeLines(
  readLines(file.path(wd, 'model', 'stan', 
                      'model8_with3predictors_nopooling_standphi_nonconcordantshock.stan'))[58:340])

```
